CROP2, a Retriever-PROPPIN Complex Mediating Protein Export from Endosomes to the Plasma Membrane

Version 1

Abstract

Endosomes generate tubulo-vesicular carriers to redistribute proteins between plasma membrane, Golgi, and lysosomes. These transport routes employ distinct combinations of sorting nexins with complexes such as Retromer or Retriever. We now show that, while Retromer associates with the PROPPIN WIPI1 to form the previously described CROP complex, Retriever associates with WIPI2, forming CROP2. WIPI2 integrates into Retriever-dependent coat complexes, since it interacts both with the Commander subunit CCDC93 and its cognate sorting nexin SNX17. CROP and CROP2 are exclusive in their physical associations and pathway selective. Whereas CROP2 is required for endosomal exit of β1-Integrin, it does not affect CROP-dependent cargos, such as EGFR or GLUT1. Vice versa, CROP is not required for β1-Integrin trafficking. WIPI1 and WIPI2 rely on similar molecular features. They use the same FSSS motif to bind to Retromer and Retriever, respectively, and an amphipathic membrane-inserting α-helix, which conveys membrane fission activity to PROPPINs. This suggests that Retromer and Retriever coats integrate distinct PROPPIN isoforms to promote fission of the respective endosomal carriers formed by them.

Introduction

Endosomal membrane traffic regulates the abundance and activity of many receptors and transporters at the plasma membrane, which renders it critical for the development and homeostasis of cells and tissues (McNally and Cullen, 2018). In line with this, perturbations in endosomal trafficking contribute to a wide range of diseases, including cancer and some of the most common neurological disorders (Carosi et al., 2023; Mellman and Yarden, 2013; Schreij et al., 2016). Many plasma membrane proteins cycle between the cell envelope, endosomes, and the Golgi. They can move on to lysosomes for degradation unless they are retrieved before endosomes fuse with lysosomes (Cullen and Steinberg, 2018). Endosomes thus serve as sorting stations at an intersection between several compartments. To collect cargo proteins, endosomes employ a variety of coats that form tubulo-vesicular carriers. Many of them are based on sorting nexins, which scaffold the membrane into narrow tubules. Sorting nexins can act on their own, such as in the ESCPE-1 complex (Kvainickas et al., 2017; Lopez-Robles et al., 2023; Simonetti et al., 2019, 2017), or together with complexes interconnecting them, such as Retromer or Retriever and the Commander complex, into which Retriever can be integrated (Bartuzi et al., 2016; Boesch et al., 2023a; Chandra et al., 2024; Healy et al., 2023; Mallam and Marcotte, 2017; McNally et al., 2017; Phillips-Krawczak et al., 2015; Seaman, 2021). Further coats, such as clathrin adapter complexes, VINE, and ACAP1-clathrin can also act on endosomes (Dai et al., 2004; Hooy et al., 2022; Li et al., 2007; Pang et al., 2014; Ren et al., 2013; Sachse et al., 2002; Shortill et al., 2022). These coat components organize a variety of endosomal protein exit routes with partially overlapping cargo specificity (Bean et al., 2017; Kvainickas et al., 2017; McGough et al., 2014; Steinberg et al., 2013).

Retromer is part of a protein coat mediating protein exit from endosomes in association with sorting nexins, such as Vps5, Vps17 or Snx3/Grd19 in yeast, and SNX27, ESCPE-1 or SNX3 in human cells (Seaman, 2021). Retromer consists of a VPS26– VPS29–VPS35 heterotrimer, which can form higher order complexes through VPS35 and VPS26 dimerization (Collins et al., 2008; Deatherage et al., 2020; Hierro et al., 2007; Kendall et al., 2022, 2020; Kovtun et al., 2018; Leneva et al., 2021; Lucas et al., 2016; Purushothaman et al., 2017). While sorting nexins themselves change the spontaneous curvature of membranes and can tubulate them (Kovtun et al., 2018; Leneva et al., 2021; van Weering et al., 2012; Zhang et al., 2021), this activity can be enhanced by the association with Retromer, which may provide additional driving force for scaffolding the membrane (Gopaldass et al., 2023).

In human cells, numerous cargos recycle back to the cell surface in a Retromer- independent manner (Steinberg et al., 2013). Many of these depend on Retriever, an evolutionarily conserved complex of VPS26C/DSCR3, VPS29 and VPS35L/C16orf62, proteins that show distant homology to the Retromer subunits (McNally et al., 2017). Whereas the overall organization and shape of Retriever resemble those of Retromer, Retriever shows different cargo specificity and distinct features. Retriever forms part of a much larger complex, Commander, which integrates Retriever with the 12-subunit CCC complex and DENND10 and cooperates with the sorting nexin SNX17 to form coats (Boesch et al., 2023a; Butkovič et al., 2024; Healy et al., 2023; Martín-González et al., 2024; Singla et al., 2024; Steinberg et al., 2012). Like Retromer, also Commander can interact with WASH and thereby link to actin (Derivery et al., 2009; Gomez and Billadeau, 2009; Guo et al., 2024; Harbour et al., 2012; Jia et al., 2012; Phillips-Krawczak et al., 2015; Seaman et al., 2009).

Cargo carriers detach from endosomes by membrane fission (Gopaldass et al., 2024; Naslavsky and Caplan, 2023). Several factors have been implicated in this detachment, among them actin and the actin-controlling WASH complex (Derivery et al., 2009; Frisby et al., 2024; Gomez and Billadeau, 2009; Harbour et al., 2012; Jia et al., 2012; Seaman et al., 2009), dynamin-like GTPases and ATPases such as Vps1 and EHD1 (Arlt et al., 2015; Chi et al., 2014; Daumke et al., 2007; Deo et al., 2018; Gokool et al., 2007; Grant et al., 2001; Solinger et al., 2020), the ESCRT-III subunit IST1 (Clippinger et al., 2024) and the PROPPIN (β-propeller that binds phosphoinositides) WIPI1 and its yeast homolog Atg18 (DeLeo et al., 2021; Gopaldass et al., 2024, 2017). Coats and actin can exert forces on the underlying lipid bilayer, which not only allow the tubule to grow but may also produce friction between lipids and protein coats that contributes to fission (Johannes et al., 2014; Simunovic et al., 2018, 2017). In addition, dynamin-like proteins actively deform the bilayer to compress and destabilize it (Antonny et al., 2016). The lipid phosphatidylinositol-(3,5)- bisphosphate (PI(3,5)P2) and/or PI5P is also required for fission (DeLeo et al., 2021; Giridharan et al., 2022; Gopaldass et al., 2017; Rivero-Ríos and Weisman, 2022; Zieger and Mayer, 2012). PROPPINs are bona-fide effectors of PI(3,5)P2 (Courtellemont et al., 2022; DeLeo et al., 2021). They associate with membranes through two phosphoinositide binding sites (Baskaran et al., 2012; Krick et al., 2012; Liang et al., 2019; Proikas-Cezanne et al., 2004; Vicinanza et al., 2015; Watanabe et al., 2012). Upon membrane binding, a hydrophobic loop, which is located between the two lipid binding sites, folds into an amphipathic α-helix that inserts into the bilayer. Thereby, PROPPINs can tubulate lipid bilayers and drive their fission (Courtellemont et al., 2022; DeLeo et al., 2021; Gopaldass et al., 2017; Gopaldass and Mayer, 2024).

PROPPINs are a protein family that binds phosphatidylinositol 3-phosphate (PI3P), phosphatidylinositol 5-phosphate (PI5P) and phosphatidylinositol 3,5-bisphosphate (PI(3,5)P2) (Baskaran et al., 2012; Busse et al., 2015; Dove et al., 2004; Gopaldass and Mayer, 2024; Jeffries et al., 2004; Vicinanza et al., 2015). They are conserved from yeast (Atg18, Atg21, and Hsv2) to humans (WIPI1, WIPI2, WIPI3/WDR45B and WIPI4/WDR45) and exist in most eukaryotes in several isoforms, which can be grouped into two evolutionary clades (Dove et al., 2004; Proikas-Cezanne et al., 2015). In mammals, Atg18, Atg21, WIPI1 and WIPI2 belong to one clade, Hsv2, WIPI3 and WIPI4 to the second (Cong et al., 2021; Polson et al., 2010). WIPI1 and WIPI2 concentrate on phagophores when autophagy is induced and they are necessary for efficient formation of autophagosomes (Dooley et al., 2014; Polson et al., 2010; Proikas-Cezanne et al., 2004). WIPI2, which has a stronger impact on autophagy, interacts with different key factors for autophagosome formation.

WIPI1 and Atg18 also support membrane fission and protein sorting on endosomes and lysosomes, but for these functions they rely on molecular features that are distinct from those important to autophagy (Courtellemont et al., 2022; DeLeo et al., 2021; Gopaldass et al., 2017; Gopaldass and Mayer, 2024). WIPI1 localizes to endo- lysosomal compartments, where it affects the size of endosomes and the exit of proteins from endosomes to lysosomes (EGF-receptor), to the Golgi (Shiga-toxin), or to the cell surface (Transferrin receptor, GLUT-1) (DeLeo et al., 2021; Jeffries et al., 2004). Based on the differential effects of substitutions in the two phosphoinositide binding sites of WIPI1 it was suggested that binding of PI3P allows WIPI1 to support the formation of endosomal transport carriers while PI(3,5)P2 and lipid binding site two are needed for membrane fission at endosomes (DeLeo et al., 2021). WIPI1 as well as its yeast homolog Atg18 associate with Retromer to form the CROP complex, which strongly augments the fission activity of the PROPPIN and makes its binding to membrane PI(3,5)P2 dependent (Courtellemont et al., 2022; Marquardt et al., 2023).

We have recently identified the Retromer- and WIPI1-containing CROP complex as an agent supporting the export of multiple cargos from endosomes, comprising both Retromer-dependent and Retromer-independent exit routes (Courtellemont et al., 2022; DeLeo et al., 2021). Since not all cargo exit from endosomes depends on WIPI1 and CROP we analysed alternative PROPPINs that might be involved in WIPI1- independent exit from endosomes. This led us to uncover a novel role for a WIPI2- Retriever complex in trafficking of β1-Integrin, which we present below.

Results

WIPI1 promotes the exit of diverse cargos from endosomes, for example for transporting Transferrin to the plasma membrane, GLUT1 to the plasma membrane, EGF-receptor (EGFR) to lysosomes, and of Shiga toxin to the Golgi (DeLeo et al., 2021). β1-Integrin provides an exception, its delivery to the cell surface being unaffected by knockout of WIPI1 (Courtellemont et al., 2022). We investigated whether the endosomal exit of β1-Integrin might depend on another PROPPIN. Since WIPI1 and WIPI2 are similar (Bakula et al., 2013; Gopaldass and Mayer, 2024), belong to the same evolutionary clade, share two predicted lipid binding sites as well as the potential to form an amphipathic helix on their CD loop 6 (Cong et al., 2021; Gopaldass et al., 2017; Polson et al., 2010), we focused our attention on WIPI2.

WIPI1 and WIPI2 are specific for distinct trafficking pathways

We tested the impact of WIPI2 knockdown (WIPI2KD) on the afore-mentioned trafficking pathways. siRNA against WIPI2 reduced the protein in whole cell extracts by 94% when compared to cells transfected with control siRNA (WIPI2CT) (Fig. S1). We analysed trafficking of EGFR from early endosomes to degradative, LAMP1- containing compartments. After 24h of serum starvation, cells were stimulated with EGF and analysed for EGFR localization by immunofluorescence. 15 min after EGF addition, EGFR-positive structures colocalized with the early endosomal marker EEA1 to more than 75% in both WIPI2KD and WIPI2WT cells, indicating that EGF had induced efficient internalization of EGFR in both cases (Figure 1). After 30 and 60 min of EGF stimulation, EGFR increasingly shifted from EEA1- towards LAMP1-positive lysosomal structures. EGFR signals declined after 60 min in the immunofluorescence images as well as in Western blot analyses of the cells (Fig. 1 C-E). Thus, WIPI2 knockdown does not impair EGFR trafficking from EEA1 endosomes to LAMP1-positive compartments and leaves lysosomal degradation of EGFR functional.

Effect of WIPI2 knockdown on EGFR degradation.

A, B) WIPI2CT and WIPI2KD cells were serum-starved for 24 h and then supplemented with EGF (100 ng/ml). After the indicated periods of time, cells were fixed, permeabilized, DAPI-labeled (blue) and decorated with antibodies to EGFR (red), EEA1 or LAMP1 (green). Scale bars: 10 μm. C) Colocalization of EGFR with EEA1 or LAMP1 was quantified over time using the images from A and B and Manders’ correlation coefficients were calculated. Values are the mean ± s.d.; (n = 3 independent experiments). 150 cells were quantified per sample. D) EGFR degradation. WIPI2CT and WIPI2KD cells were stimulated with EGF for different periods of time, lysed and subjected to SDS-PAGE and Western blot analysis for EGFR. α-Tubulin served as a loading control. E) Quantification of EGFR from (E), using the value at time 0 (cells starved for 24 h) as 100% reference. Data are means ±s. d., from three independent experiments.

To test the role of WIPI2 in Retromer-dependent trafficking we assayed the localisation of the glucose importer GLUT-1 because this protein accumulates on endosomes if its recycling to the plasma membrane through Retromer is impaired (Steinberg et al., 2013). To quantify the surface fluorescence, we acquired z-stacks of non- permeabilized, immunostained cells, generated maximum intensity projections and integrated the fluorescence signal over the entire cell. Non-permeabilized WIPI2KD cells showed similar cell surface staining of GLUT-1 as WIPI2CT cells (Figure 2 A-B), suggesting that also Retromer-mediated recycling of GLUT1 occurs independently of WIPI2.

GLUT1 expression and localization upon WIPI2 knockdown.

A. GLUT1 cell surface exposure. Control and WIPI2KD cells were fixed and stained with antibody against GLUT1 and with DAPI. Where indicated, cells had been permeabilized with 0.05% before staining. Scale bars: 10 μm. B. Quantification of GLUT1-immunofluorescence in cells from A. Regions of interest (ROIs) corresponding to each cell and in some regions outside the cells (background) were manually defined using image J software. Total cell fluorescence was integrated and corrected for background fluorescence. 150 cells per condition, stemming from three independent experiments, were quantified. Red bars show the means. P values were calculated applying an unpaired Student’s t-test with unequal variances. The analysis was performed with 99% confidence. NS = not significant (P > 0.05).

Next, we investigated the requirement of WIPI2 in the trafficking of β1-Integrin. Surface accumulation of this Integrin subunit is unperturbed by loss of SNX27 or Retromer but depends on the SNX17-Retriever pathway (McNally et al., 2017; Steinberg et al., 2013). Knockdown of WIPI2 induced a loss of β1-Integrin immunofluorescence from the surface of HK2 cells. The effect was comparable to that of a knockdown of SNX17 (Figure 3 A, B). When the plasma membrane of the cells was detergent-permeabilized to improve accessibility of internal structures for immunostaining, cells depleted of WIPI2 showed an accumulation of β1-Integrin on intracellular structures. This signal colocalized with the early endosome marker EEA1 and, to a lesser degree, with the late endosomal/lysosomal marker LAMP1 (Figure 3 C-D).

Impact of WIPI2 knockdown on β1-Integrin localization.

A. β1-Integrin cell surface expression. Control and WIPI2KD cells were fixed and stained with antibody to β1- Integrin (green) and with DAPI (blue), without detergent permeabilization. Scale bars: 10 mm. B. Quantification of β1-Integrin immunofluorescence in cells from A. Regions of interest (ROIs) corresponding to each cell and in some regions outside the cells (background) were manually defined using ImageJ software. Total cell fluorescence was integrated and corrected for background fluorescence. 150 cells per condition, stemming from three independent experiments, were quantified. Red bars show the means. P values were calculated applying unpaired Student’s t-test with unequal variances. The analysis was performed with 99% confidence. ***P < 0.001. C. Immunofluorescence staining of intracellular β1-Integrin (green) and EEA1 (red) or LAMP1 (red) in CTRL and WIPI2KD cells. Cells were fixed, permeabilized with 0.05% saponin and stained. Scale bars: 10 mm. D. Colocalization of β1-Integrin with LAMP1 or EEA1 was measured in cells from C, using Manders’ colocalization coefficient M2, calculated in Image J. Colocalization was quantified from three independent experiments with a total of 150 cells. An unpaired Student’s t-test with unequal variances was used to calculate P-values. The analysis was performed with 99% confidence. ***P < 0.001.

In accord with previous observations (DeLeo et al., 2021), β1-Integrin surface labelling in non-permeabilized cells was not affected by knockdown of WIPI1 (Figure S2). Thus, β1-Integrin recycling to the plasma membrane selectively requires WIPI2.

The conserved amphipathic α-helical and FSSS motifs of WIPI2 promote β1- Integrin recycling

WIPI2 shares two critical features with other PROPPINs (Gopaldass and Mayer, 2024). These include an FSSS motif in the membrane-distal part and the CD loop on blade 6 of the β-propeller structure. The CD loop promotes membrane fission activity of Atg18 during the fragmentation of yeast vacuoles (Gopaldass et al., 2017) and in the detachment of endosomal carriers (DeLeo et al., 2021). Upon contact with the bilayer, the loop folds into an amphipathic α-helix that is conserved among PROPPINs and necessary for membrane fission activity. To test whether the amphipathic nature of this helix also contributes to the function of WIPI2 in β1-Integrin retrieval to the cell surface, we swapped two pairs of amino acids on opposite sides of this α-helix in WIPI2 (Figure 4 A). This manipulation preserves the amino acid composition of the helix but abolishes its amphipathic character. The resulting EGFPWIPI2SLoop (scrambled loop) variant and the wildtype version EGFPWIPI2WT were expressed in WIPI2CT and WIPI2KD cells at comparable levels, as shown by Western blot (Figure S3). Whereas permeabilized WIPI2KD cells displayed β1-Integrin concentrated on perinuclear endosomes, expression of EGFPWIPI2WT rescued this phenotype (Figure 4 B). EGFPWIPI2Sloop failed to induce such recovery in WIPI2KD cells. It even exerted a dominant-negative effect when expressed in WIPI2CT cells. (Figure 4 C-D). Cells expressing EGFPWIPI2SLoop also showed conspicuous long tubules, which were more readily detectable in live cell imaging (Figure S4) than in fixed cells (Figure 4).

Role of the amphipathic α-helix of WIPI2 in β1-Integrin sorting.

A. The amphipathic α-helix. Helical wheel projections showing the CD-loop on blade 6 of the wildtype sequence, WIPI2WT, and WIPI2Sloop. Coloured arrows indicate the two pairs of amino acids that have been swapped in WIPI2Sloop. The magnitude and direction of the hydrophobic moment of the helices was predicted using the online tool Heliquest (Gautier et al., 2008). It is indicated by the vector in the centre of the wheels. Sequences of the hydrophobic loop region of WIPI2WT and WIPI2Sloop are shown, and predicted α-helices are plotted in red. The two pairs of hydrophobic/hydrophilic amino acids that are swapped in WIPI2Sloop are highlighted by rectangles in the sequences. B,C. β1-Integrin localisation. WIPI2KD (B) and WIPI2CT (C) cells were transfected with a plasmid carrying siRNA-resistant EGFPWIPI2WT or EGFPWIPI2Sloop. After 18h of viral transfection, cells were fixed, permeabilized with 0.05% saponin and stained with DAPI and antibodies to β1-Integrin. Scale bars: 10 mm. D. Quantification of β1-Integrin immunofluorescence in cells from B and C. Regions of interest (ROIs) corresponding to cells expressing the indicated WIPI2 variants, and some regions outside the cells (background), were manually defined using Image J software. Total cell fluorescence was integrated and corrected for background fluorescence. 105 cells per condition stemming from three independent experiments were analysed. Red bars indicate the means. P-values were calculated by Welch’s t-test. The analysis was performed with 99% confidence: ****P < 0.00001. NS: not significant (P>0.05)

Alignments of Atg18 and WIPI1/2 orthologs have revealed an FSS/TS motif in blade 2 (Gopaldass et al., 2017; Gopaldass and Mayer, 2024). On the β-propeller structure of these PROPPINS, this motif is located at the membrane-distal side. It is a potential phosphorylation site and required for the interaction of Retromer with WIPI1 and Atg18 (Courtellemont et al., 2022). Since the overall structural arrangement of Retriever resembles that of Retromer (Bartuzi et al., 2016; Boesch et al., 2023a; Healy et al., 2023; McNally et al., 2017), we tested whether WIPI2 requires its FSSS motif to bind Retriever. To this end, we generated WIPI2S67E and WIPI2S67A to substitute the central serine in the FSSS motif and mimic effects of potential phosphorylation at this site.

The functionality of these WIPI2 variants was tested by rescuing WIPI2KD cells through expression of siRNA resistant variants of EGFPWIPI2. Both EGFPWIPI2S67E and EGFPWIPI2S67A were expressed to similar levels as an EGFPWIPI2 wildtype allele (Figure S3). Immunofluorescence staining of non-permeabilized cells showed that the expression of EGFPWIPI2WT re-established the exposure of β1-Integrin at the cell surface, while EGFPWIPI2S67A and EGFPWIPI2S67E were ineffective (Figure 5 A-C). Expression of the two variants also impaired surface exposure of β1-Integrin in WIPI2CT cells (Figure 5 B-D). This dominant negative effect is significant because it implies that the two variants compete with the endogenous protein. To be able to do so they should be folded, suggesting that the S67 substitutions do not grossly perturb WIPI2 structure. A conspicuous consequence of both of both FSSS substitutions was the accumulation of EGFPWIPI2S67A and EGFPWIPI2S67E in long tubules, which often interconnected several dot-like structures, presumably endosomes (Figs. 5 and S4). This suggests that both alleles might permit the formation of tubulo-vesicular endosomal structures but interfere with their detachment. Taken together, these observations support the notion that WIPI2 contributes fission activity to Retriever- mediated trafficking through its amphipathic helix and its FSSS motif.

Effects of the WIPI2 FSSS motif on β1-Integrin recycling.

A,B. Influence of WIPI2 variants on β1-Integrin. WIPI2-knockdown (WIPI2KD, A) and control (WIPI2CT) HK2 cells were transfected with a plasmid carrying siRNA-resistant wildtype or the indicated FSSS variants of EGFPWIPI2. After 18 h of viral transfection, cells were fixed and stained (without detergent permeabilisation) with DAPI and antibodies to β1-Integrin (red). Scale bars: 10 mm. C,D. Quantification of β1-Integrin immunofluorescence in cells from A and B. Regions of interest (ROIs) corresponding to cells expressing the indicated WIPI2 variants, and some regions outside the cells (background), were manually defined using Image J software. Total cell fluorescence was integrated and corrected for background fluorescence. 105 cells per condition stemming from three independent experiments were analysed. Red bars indicate the means. P-values were calculated by Welch’s t-test. The analysis was performed with 99% confidence: ****P < 0.00001.

WIPI2 interacts with Retriever, Commander and SNX17

Next, we tested whether WIPI1 and WIPI2 physically interacts with Retriever or Retromer and could become part of their respective coats. In that case, WIPI2 should interact not only with Retriever itself, but also with the sorting nexin SNX17 and perhaps the CCC subunits, which integrate Retriever into the Commander complex (Butkovič et al., 2024; Martín-González et al., 2024; Singla et al., 2024). We assayed this by co-immunoadsorption. An HA-tagged allele (WIPI2HA) was expressed in HK2 cells using lentiviruses. Extracted proteins were adsorbed to anti-HA beads and analysed by Western blotting (Figure 6). Whereas the Commander subunit CCDC93 and SNX17 co-adsorbed to the beads in extracts from cells expressing WIPI2HA (Figure 6 A), these signals were absent from cells expressing only non-tagged WIPI2. Co-adsorption of SNX17 was abolished by knockdown of the Retriever subunit VPS26C (Figure 6B). Although VPS26C is quite similar to the Retromer subunit VPS26, only VPS26C co-adsorbed with WIPI2HA but VPS26 was not detectable (Figure 6 C-D). The situation was inversed when the HA tag was attached to WIPI1. In this case, VPS26 was recovered on the beads but VPS26C was not (Figure 7A, B). SNX27, a sorting nexin that interacts with Retromer (Simonetti et al., 2022; Temkin et al., 2011), was also co-adsorbed with WIPI1HA (Figure 7C). This suggests that Retriever bridges WIPI2 and SNX17 and that Retromer bridges WIPI1 with SNX27. These PROPPINs might thus be selectively integrated into the respective cognate Retriever- or Retromer- coats.

Integration of WIPI2 with coat subunits.

A. Interaction of WIPI2HA with CCDC93. HK2 cells and HK2 cells stably expressing WIPI2HA were detergent solubilised, the total cell extracts were incubated with anti-HA beads and washed. Adsorbed protein was analysed by SDS-PAGE and Western blotting using the indicated antibodies. The intensity of the interacting CCDC93 was quantified with a LICOR Odyssey fluorescence imager. The background from the corresponding position in the sample from cells without HA-tag was subtracted. The resulting intensity is shown relative to the intensity of WIPI2HA signal on the beads. N = 3 biological replicates. Red bars show the means and error bars represent the SEM. P values were calculated with an unpaired Student’s t-test. *P < 0.01. B. Interaction of WIPI2HA with SNX17. HK2 cells stably expressing WIPI2HA were treated with siRNA to VPS26C (VPS26KD) or with non-specific siRNA (VPS26CT) and lysed. The total cell extracts were incubated with anti-HA beads and adsorbed proteins were analysed using the indicated antibodies as in A. N = 3. C,D. Selectivity for Retriever versus Retromer. The co-immunoadsorption experiments were performed as in A and analysed for co-adsorbed (C) VPS26C or (D) VPS26. N=3. *P < 0.01.**P < 0.001. NS=not significant.

Interaction of WIPI1 with VPS26 and SNX27, but not VPS26C.

A. Interaction of WIPI1HA with VPS26. HK2 cells and HK2 cells stably expressing WIPI1HA were detergent solubilised, the total cell extracts were incubated with anti-HA beads and washed. Adsorbed protein was analysed by SDS-PAGE and Western blotting using the indicated antibodies. The intensity of the interacting VPS26 was quantified with a LICOR Odyssey fluorescence imager. The background from the corresponding position in the sample from cells without HA-tag was subtracted. The resulting intensity is shown relative to the intensity of WIPI1HA signal on the beads. N = 3 biological replicates. Red bars show the means and error bars represent the SEM. P values were calculated with an unpaired Student’s t-test. B. Lack of interaction of WIPI1HA with VPS26C. HK2 cells stably expressing WIPI1HA were used for co-immunoadsorption experiments as in A and decorated with the indicated antibodies. N = 3. C. Interaction of WIPI1HA with SNX27. HK2 cells stably expressing WIPI1HA were used for co-immunoadsorption experiments as in A and decorated with the indicated antibodies. N = 3. **P < 0.001. NS: not significant.

Since the FSSS motif of WIPI2 is necessary for its transport functions, we tested whether this motif affects the interactions of WIPI2 with Retriever. To this end, wildtype and the two FSSS variants of WIPI2HA were stably expressed and cell extracts were probed by the same co-adsorption approach as above (Fig. 8). Whereas the Retriever subunit VPS26C was efficiently co-adsorbed with WIPI2WT-HA, WIPI2S67A-HA and WIPI2S67E-HA labilized this interaction. Cells carrying only the non-tagged, endogenous WIPI2 yielded no VPS26C signal. Thus, the FSSS motif is necessary for the interaction between WIPI2 and Retriever.

Impact of the FSSS motif on the WIPI2-Retriever interaction.

A. HK2 cells and HK2 cells stably expressing the indicated WIPI2HA variants were detergent solubilised. Anti-HA beads were incubated with the total cell extracts, washed, and adsorbed proteins were analysed by SDS-PAGE and Western blotting using the indicated antibodies. B. Band intensities from the blots in A were quantified with a LICOR Odyssey infrared fluorescence imager and plotted as the ratio of VPS26C over WIPI2HA. N=3 independent biological replicates. Red bars show the means and error bars represent the SEM. P values were calculated with an unpaired Student’s t-test. **P < 0.001. HC: Heavy chain.

The interaction of WIPI2 and Retriever is necessary for their localization on Rab11-positive recycling endosomes

To test for the in vivo consequences of the WIPI2-Retriever interaction, we analysed the effect of the FSSS motif on the colocalization of the respective WIPI2EGFP variants with different endosomal RAB-GTPases, which were N-terminally tagged with mCherry. In line with previous studies (Polson et al., 2010; Puri et al., 2018) EGFPWIPI2WT mostly co-localized with mCherryRAB11 (58%) (Figures 9 and S5-S7) and much less with mCherryRAB5 (23%) and mCherryRAB7 (6%). By contrast, EGFPWIPI2S67A (63%) and EGFPWIPI2S67E (60%) co-localized mainly with mCherryRAB5 (60-63%) and mCherryRAB7 (36-38%), and much less with mCherryRAB11 (12-14%). The Retriever subunit VPS35L was detected by immuno-staining. It co-localized mainly with mCherryRAB11 in cells expressing WIPI2WT (42%) and less with mCherryRAB5 (31%) and mCherryRAB7 (23%) (Figures. 9, S5-S7). In EGFPWIPI2S67E and EGFPWIPI2S67A expressing cells, VPS35L distribution shifted, similarly as EGFPWIPI2, from mCherryRAB11 compartments (12-13%) towards mCherryRAB5 (42-44%) and mCherryRAB7 compartments (26-27%). Collectively, these results suggest that the FSSS motif in blade 2 of WIPI2 allows this PROPPIN to associate with Retriever, and that this association promotes re-distribution of WIPI2 and Retriever from Rab5 and Rab7 endosomes to Rab11 compartments.

Role of the WIPI2 FSSS motif for recruiting WIPI2 and Retriever to Rab11 endosomes.

WIPI2-knockdown HK2 cells were transfected with plasmids expressing the indicated siRNA- resistant EGFPWIPI2 variants and mCherry-RAB5, mCherry-RAB7, or DsRed-Rab11 were fixed, permeabilized, immuno-stained for VPS35L and analyzed by confocal microscopy. Examples of the quantified images are presented in Supplementary Figures S5 to S7. Colocalization with EGFPWIPI2 and VPS35L was assessed for: A. RAB11 B. RAB5 C. RAB7. Colocalization was quantified using Manders’ colocalization coefficient M2, calculated in ImageJ. M2 refers to the fraction of VPS35L colocalising with the different Rab-proteins. For the triple colocalizations WIPI2/VPS35L/RAB M2 indicates the fraction of green pixels (WIPI2) overlapping with the pixels positive for the VPS35L/RAB colocalization. Colocalization was quantified from three independent experiments in a total of 120 cells. P values were calculated applying Welch’s t-test with unequal variances. The analysis was performed with 99% confidence. ***P < 0.0001.

Discussion

Our experiments reveal a pathway specificity for WIPI1 and WIPI2. The two proteins support distinct cargoes and transport routes and interact with different partners.

This isoform selectivity contrasts with the roles of these proteins in their second functional domain, autophagy, which strongly depends on WIPI2 but is also significantly reduced by ablation of WIPI1 (DeLeo et al., 2021; Dooley et al., 2014; Guan et al., 2001; Obara et al., 2008; Polson et al., 2010; Proikas-Cezanne et al., 2015, 2004). Furthermore, the autophagic function of the PROPPINs also exploits other molecular features than endosomal protein exit. For autophagy, WIPI1 and Atg18 do not require the amphipathic helix in CD loop 6, which is essential for endosomal transport. It was hence proposed that these PROPPINs perform distinct functions in autophagy and in endosomal protein exit, and that their membrane fission activity may only be essential for endosomal transport (Courtellemont et al., 2022; DeLeo et al., 2021; Gopaldass et al., 2017).

In endosomal transport, WIPI1 and WIPI2 display selectivity in their partners. WIPI1 interacts with Retromer and WIPI2 with Retriever, mirroring the impact of the two PROPPINs on the respective transport routes and cargos that they support. WIPI1 and WIPI2 utilise the same molecular features to support these transport routes.

Their FSSS motif is necessary for both isoforms to integrate with Retromer and Retriever, respectively. This, together with the similarity in the overall organisation of Retromer and Retriever, suggest that they might form complexes in a similar manner. In analogy to the WIPI1-Retromer complex, which was termed CROP (Courtellemont et al., 2022), we hence refer to the WIPI2-Retriever complex as CROP2.

WIPI2-dependent transport of β1-Integrin depends also on the amphipathic character of the α-helix in CD loop 6. In Atg18 and WIPI1, this helix is necessary to convey membrane fission activity, which is used both for division of the lysosome-like vacuoles of yeast and for detachment of endosomal carriers (DeLeo et al., 2021; Gopaldass et al., 2017; Zieger and Mayer, 2012). This fission activity is potentiated by integrating Atg18 and WIPI1 with Retromer into the CROP complex (Courtellemont et al., 2022). Our observations suggest that WIPI2 contributes pathway-specific membrane fission activity by integrating with Retriever. In line with this, ablating the amphipathic character of its helix on CD loop 6 disrupts β1-Integrin recycling to the cell surface and leads to its accumulation in endosomal compartments. Disrupting the WIPI2-Retriver interaction by substitution of the FSSS motif has the same effect. It also leads to the accumulation of long tubular structures, which we presume to be exaggerated endosomal carriers that continue to grow but fail to detach. This phenotype is consistent with a block in the membrane fission step that is necessary to allow the endosomal carrier to depart.

Retriever couples to other proteins that are necessary for endosomal protein transport. These include the CCC complex, which associates with Retriever to yield the very large Commander complex (Boesch et al., 2023b; Healy et al., 2023; Laulumaa et al., 2024). How Commander is placed on the membrane is currently unknown, but the interaction between SNX17 and Retriever contributes to its recruitment (Butkovič et al., 2024; Healy et al., 2022; Singla et al., 2024; Steinberg et al., 2012) and presumably to the formation of a coat. Our pull-downs of WIPI2 detected not only the Retriever subunit VPS26C as an interactor, but also SNX17 and the CCC subunit CCDC93. This suggests that WIPI2 integrates into the coat and might thus bring membrane fission activity to the place where it is required for carrier formation. Here, it could synergize with other factors that support membrane fission in endosomal transport, such as EHD1, which interacts with SNX17, and the actin- regulating WASH complex, which binds to Commander (Bartuzi et al., 2016; Dhawan et al., 2022, 2020; Phillips-Krawczak et al., 2015).

The requirement of two distinct WIPI proteins, using the same molecular features, in Retromer- and Retriever-dependent transport uncovers a novel conserved element between these two pathways. In combination with the similar overall organisation of Retromer and Retriever and their common interactions with EHD proteins (Arlt et al., 2015; Chi et al., 2014; Daumke et al., 2007; Deo et al., 2018; Gokool et al., 2007; Grant et al., 2001; Solinger et al., 2020) and WASH (Derivery et al., 2009; Gomez and Billadeau, 2009; Guo et al., 2024; Harbour et al., 2012; Jia et al., 2012; Phillips-Krawczak et al., 2015; Seaman et al., 2009), this supports the notion that these two membrane coats share similar mechanistic principles for their formation and/or detachment.

Materials and Methods

Antibodies and reagents

All chemical reagents were from Sigma-Aldrich unless specified otherwise

We used anti-LAMP1 (H4A3, US Biological, Life Sciences); anti- EEA1 (#2411, Cell Signaling); anti-EGFR (sc-373746, clone A-10 Santa Cruz Biotechnology; PA5- 85089 ThermoFisher Scientific); anti-Integrin β1 ( ab150361, Abcam); anti-GLUT1 (ab15309, Abcam); anti-c16orf62/VPS35L (PA5-28553, ThermoFisher Scientific); anti-SNX17(HPA043867, Atlas Antibodies); anti-SNX27 (ab77799, Abcam); anti CCDC93 (20861-1AP, Proteintech); anti-α-Tubulin (T9026, Sigma); anti-Vps26 (ab181352, Abcam); anti-VPS26C/DSCR3 (ABN87, Merck Millipore), WIPI1 (W2394, Sigma); WIPI2 (HPA019852, Sigma).

As secondary antibody we used: Cy3-conjugated Affinity Pure Donkey anti-Mouse IgG H + L (Jackson Immuno Research); Cy3-conjugated AffiniPure Donkey anti- Rabbit IgG H + L (Jackson Immuno Research); Alexa Fluor®488-conjugated AffiniPure Donkey anti-Rabbit IgG H + L (Jackson Immuno Research); Alexa Fluor®488-conjugated AffiniPure Donkey anti-Mouse IgG H + L (Jackson Immuno Research); Alexa Fluor® 647-conjugated AffiniPure donkey anti-rabbit IgG H + L (Jackson Immuno Research); Alexa Fluor® 647-conjugated AffiniPure donkey anti- Mouse IgG H + L (Jackson Immuno Research). Antibodies were used at dilutions summarized in Supplementary Table1.

Other reagents: Opti-MEM (Thermo Fischer, 11058021) and Trypsin (Thermo Fischer, 27250018).

Protease inhibitor (PI) cocktail final concentrations in samples: 100 μM pefablock SC (Merck, 11429876001), 2 μM leupeptin (Merck, 11529048001), 50 μM o-phenanthroline (Merck ,131377), 0.7 μM pepstatin A (Merck, 11524488001) dissolved in methanol.

Phosphatase inhibitor cocktail final concentrations: 10 mM NaF (S-7920, Sigma); 1mM Na Orthovanadate (S6508, Sigma); 2 mM β-glycerophosphate (A2253, AppliChem) all dissolved in water.

Complementary DNA (cDNA) constructs

Vectors expressing tagged RAB proteins were purchased from Addgene: DsRed- RAB11 (12679; deposited by R. Pagano); mCherry-RAB7 (61804; deposited by G. Voeltz); mCherry-RAB5 (49201; deposited by G. Voeltz). The following vectors were kindly provided by colleagues: EGFP-WIPI2 (pAR31CD vector) (Tassula Proikas- Cezanne, Tübingen, Germany). The sequences of WIPI1 and WIPI2 that have been used are given in Supplementary File 1.

Site-directed mutagenesis

All constructs were verified by DNA sequencing. EGFPWIPI2 or WIPI2HA was used as DNA template for site-directed mutagenesis (QuikChange mutagenesis system, Agilent Technologies, 200524) to generate point mutations in the FSSS motif (S67A and S67E) following the manufacturer’s protocol.

To get WIPI2S67E we used the following primers: Fw: AGATTGTTCTCCGAGAGCCTAGTGGCC and Rv GGC CAC TAG GCT CTC GGA

GAA CAA TCT; while for WIPI2S67A we used Fw: AGATTGTTCTCCGCTAGCCTAGTGGCC

Rv: GGCCACTAGGCTAGCGGAGAACAATCT (Microsynth).

WIPI2SLoop was generated by four consecutive amplifications using the following primers:

Fw: TGGACC GGGTACAAAGGGAAAGTG CTC Rv: GAGCACTTTCCCTTTGTACCCGGTCCA;

Fw: GGGTACAAAGGGTTCGTGCTCATGGCC Rv: GGCCATGAGCACGAACCCTTTGTACCC;

Fw: CTC ATG GCC TCC TAC AGC TAC CTG CCT Rv: AGGCAGGTAGCTGTAGGAGGCCATGAG;

Fw: GCC TCC TAC AGC ACC CTG CCT TCC CAA Rv: TTGGGAAGGCAGGGTGCTGTAGGAGGC

Non-mutated template vector was removed from the PCR mixture through digestion by the enzyme Dpn1 for 1 h at 37°C. The product was purified using NucleoSpin PCR Clean-up (Macherey-Nagel, 740609.50S) and transformed into Escherichia coli. Plasmid DNA was purified and sequenced. The site-directed mutagenesis system was also used to generate the siRNA-resistant cDNA for WIPI2 with primers Fw: CGATAGTCCTTTAGCCGCA and Rv: TGCGGCTAAAGGACTATCG. All oligonucleotides were synthesized by Microsynth.

Cell culture, transfection and treatments

HK2 cells were grown in DMEM-HAM’s F12 (Thermo Fischer, 11765054) supplemented with 5% fetal calf serum (Gibco, 10270106), 50 U/mL penicillin/50 mg/mL streptomycin (ThermoFischer, 15140148), ITS (5 μg/mL insulin, 5 μg/mL TF, 5 ng/mL selenium; LuBio Science, 00-101-100ML). Cells were grown at 37°C in 5% CO2 and at 98% humidity.

HK2 cells were transfected with plasmids using the X-tremeGENETM HP DNA transfection reagent (Sigma-Aldrich, 6366546001), unless otherwise specified, according to the manufacturer’s instructions, and incubated for 18-24 h before fixation or live cell imaging. The HK2 cell line was checked for mycoplasma contamination by a PCR-based method. All cell-based experiments were independently repeated at least three times.

Generation of HK2 cells expressing WIPI2HA

HK2 cell lines stably expressing the three forms (WT, S67A, S67E) of C-terminal HA- tagged WIPI2 were generated by lentiviral transduction. WIPI2 was cloned into modified pLKO.1 lentiviral vector (available from Sigma) between Xba1 and EcoR1 restriction sites. To obtain WIPI2HA a sequence of 3Gly - 1Ser - 3Gly - 1Ser - HA - HA (designed and purchased from GeneScript) was added using Gibson assembly according to the manufacturer’s protocol. Finally, the lentiviral plasmid DNA was purified and sequenced. Lentiviral transfer vector together with third generation envelope and packaging plasmids were transfected into 293T cells to generate lentivirus. Packaged lentivirus containing supernatant was added to recipient HK2 cells in the presence of 10 μg/ml Polybrene (AL-118, Sigma-Aldrich). HK2 cells were stably infected with a lentiviral vector for constitutive WIPI2HA expression for 6 h. 2 days after infection, cells were selected with 1 μg/ml of puromycin for 3 days. Protein samples were collected and used for IP experiments 7 days after infection.

RNA interference

HK2 cells were transfected with siRNA for 72 h using Lipofectamine® RNAiMax (Thermo Fisher Scientific, 13778150) according to the manufacturer’s instructions. Control cells were treated with identical concentrations of siGENOME Control Pool Non-Targeting from Dharmacon (D-001206-13-05). siRNAs targeting WIPI2 were from Dharmacon (ON-TARGETplus Human WIPI2 J-020521). siRNAs were used at a final concentration of 20 nM.

Endocytosis assays

EGFR degradation

HK2 cells were seeded in DMEM-HAM’s F12 supplemented with 5% foetal calf serum, 50 U/mL penicillin/50 mg/mL streptomycin and ITS (5 μg/mL insulin, 5 μg/mL TF, 5 ng/mL selenium. The day after the growth medium was replaced with the same but without serum for 24 h and HK2 cells were stimulated with 100 ng/ml EGF for the indicated periods of time. The cells were fixed for 10 min in 4% paraformaldehyde in PBS at different time points after stimulation.

Immunofluorescence

Cells were fixed for 10 min in 4% paraformaldehyde in PBS. After fixation, cells were permeabilized in 0.05% (w:v) saponin (Sigma-Aldrich, 558255), 0.5% (w:v) BSA and 50 mM NH4Cl in PBS (blocking buffer) for 30 min at room temperature. The cells were incubated for 1 h with primary antibodies in blocking buffer, washed three times in PBS, incubated for 1 h with the secondary antibodies (Alexa Fluor-conjugated), washed three times in PBS, mounted with fluorescence mounting medium (Dako, S3023) on slides and analysed by confocal microscopy.

Confocal fluorescence microscopy, image processing, and colocalization analysis

HK2 cells were grown to 70% confluence on glass coverslips and immunofluorescence microscopy was performed as described above. The experiments were repeated at least three times and representative images are shown. Colocalization was analysed by acquiring serial sections from about 100 cells per sample. Images were exported in TIFF format and processed as previously described (Vicinanza et al., 2011). Images of samples to be compared were acquired using the same settings (i.e. laser power, photomultiplier gain and pinhole size), avoiding pixel saturation. The images were processed in the same way using ImageJ software. Channels from each image were converted into 8-bit format and the ‘Auto Local Threshold’ Plug-in with the ‘Default’ method was used to segment grayscale images, identify the structures of interest and subtract background.

GLUT1 and β1-Integrin fluorescence levels were quantified in CTR and WIPI2 KD cells using ImageJ. z-stack images were acquired and compressed into a single plane using the ‘maximum intensity Z-projection’ function in Image J. Individual cells were selected using the freeform drawing tool to create a ROI (ROI). The ‘Measure’ function provided the area, the mean grey value and integrated intensity of the ROI. The mean background level was obtained by measuring the intensity in three different regions outside the cells, dividing them by the area of the regions measured, and averaging the values obtained. The background was subtracted from each cell to obtain the CTCF (corrected total cell fluorescence). It was calculated using the formula: CTCF=integrated intensity of cell ROI − (area of ROI × mean fluorescence of background).

To quantify the degree of colocalization, confocal z-stacks were acquired, single channels from each image in 8-bit format were thresholded to subtract background and then the "Just Another Colocalisation Plug-in" (JACOP) was used to measure the overlap coefficient according to Manders. Manders’ coefficient indicates the overlap of the signals and represents the degree of colocalization: where S1 represents signal intensity of pixels in the channel 1 and S2 represents signal intensity of pixels in the channel 2. The colocalization coefficients M1 and M2 indicate the contribution of each channel to the pixels. They are independent of the intensity of the overlapping pixels, but they are sensitive to background, so a threshold must be set (Manders et al., 2011). In our colocalization analysis, M1 indicates the fraction of red pixels overlapping with green one, while M2 indicates the fraction of green pixels overlapping with the red ones. Confocal microscopy was performed on an inverted confocal laser microscope (Zeiss LSM 880 with airyscan) with a 63x 1.4 NA oil immersion lens unless stated otherwise.

Gel electrophoresis and Western blot

Cells were plated into 12-well tissue culture test plates (TPP) until they reached 90% confluency, except for knockdown-cells, which were cultured until 72h after transfection with the siRNAs. Cells were then washed three times with ice-cold PBS (phosphate-buffered saline), scraped, and proteins were extracted in ice-cold lysis buffer (150 mM NaCl, 2 mM EDTA, 40 mM HEPES-NaOH pH 7.4, and 1% Triton X- 100 supplemented with phosphatases and protease inhibitor cocktail (see details in Reagent section). Protein extracts were supplemented with 1/4 volume of 5x reducing sample buffer (250 mM Tris-Cl, pH 6.8, 5% β-mercaptoethanol, 10% SDS, 30% glycerol, 0.02% bromophenol blue) and heated to 95 °C for 5 min. The samples were run on either 8%, 10% or 12.5% SDS-polyacrylamide gels (W x L x H: 8.6 x 6.8 x 0.15 cm). The stacking gels were prepared as follows: 6% acrylamide, 0.16% bis- acrylamide, 0.1 M Tris, pH 6.8, 0.1% SDS, 0.1% TEMED, 0.05% ammonium persulfate. Running gels were: 10% or 12.5% acrylamide, 0.27% or 0.34% bis-acrylamide, 0.38 M Tris, pH 8.8, 0.1% SDS (Applichem, 475904-M), 0.06% TEMED (Applichem, A1148), 0.06% APS (Applichem, A2941). The gels were run at constant current (20–30 mA). Proteins were blotted onto nitrocellulose membrane by the semi-dry method for 80 min at 400 mA (Trans-Blot® SD Semi-Dry Electrophoretic Transfer Cell, Bio-Rad). After incubation with the primary antibody, signals were detected by secondary antibodies coupled to infrared dyes (LI-COR) and detected on a LI-COR Odyssey infrared fluorescence imager. Images were exported as TIFF files and processed in Adobe Photoshop. Band intensity was quantified using ImageJ band analysis.

Immunoprecipitation of Retriever

HK2 cells expressing the three variants (WT, S67E, S67A) of WIPI2 tagged with HA were grown to 80% confluence in 15 cm dishes. After rinsing the cells three times with phosphate buffered saline (PBS 1X), they were detached and lysed by incubating 15 min at 4°C with lysis buffer: 150 mM NaCl, 2 mM EDTA, 40 mM HEPES-NaOH pH 7.4, 1% Triton X-100, supplemented with phosphatases and proteases inhibitor cocktails (details in Reagent section) and 1 mM PMSF out of a 200 mM solution in ethanol, which was freshly prepared immediately before use. The lysate was spun for 10 min at 10,000 x g on a cold bench-top centrifuge). 2.5% of the clarified extracts was reserved as “input”. Each clarified extract was incubated for 2 h at 4°C with 20 μl of Anti-HA High Affinity antibody (clone 3F10) (Roche #118150116001). Then, samples were washed three times with lysis buffer and, after discarding the last wash, the beads were resuspended in 50 μl of pre-warmed elution buffer (2% SDS, 1mM EDTA, 10mM DTT, 20mM Tris HCl pH 7.4, 100 µM PMSF) and shaken at 55°C for 10 min. Samples were centrifuged, the supernatants were transferred to new Eppendorf tubes, and 5x sample buffer containing 100 mM of fresh DTT was added. Samples were kept at 95°C for 5 min and centrifuged in a table to centrifuge. All supernatants were transferred to a new Eppendorf tube yielding the “IP” samples. Inputs and IP were loaded on 8% or 10% SDS– polyacrylamide gels.

Statistical analysis

Differences between the means have been evaluated by two-sample Student’s t-test assuming unequal variances unless otherwise specified. They are mentioned as significant differences when p < 0.01. Immunofluorescence experiments were independently repeated three times and at least 35 cells were analysed from a single experiment. Representative images are shown. Western blotting experiments were independently repeated at least three times and representative blots are shown. Most data are presented as the means ± standard deviation (s.d.) unless otherwise specified.

eLife Assessment

The authors present evidence for a WIPI2-Retriever complex (termed CROP2) that couples cargo selection to carrier fission at endosomes. CROP2 appears to function analogously to the previously described CROP1 complex, formed by WIPI1 and Retromer, with which it shares structural similarities. They provide convincing evidence that CROP1 and CROP2 regulate the trafficking of distinct subsets of cargoes; however, the cellular evidence for the existence of these distinct complexes remains incomplete. Overall, the findings are important and expand our understanding of how cargo selection by Retriever and Retromer is orchestrated at endosomes.

Reviewer #1 (Public review):

WIPI1 is a PROPPIN family protein that has been implicated in Retromer-mediated membrane fission events. Although the cargos that it has been tested to be important for are diverse, one of the cargos that is unaffected is Beta1-Integrin. This leads the authors to assess another PROPPIN family protein - WIPI2, which is a homolog of WIPI1. KD using siRNA is effective and had no consequences on LAMP1, EGFR trafficking or GLUT1 trafficking. Integrin-B1, however, had a large and significant defect in its recycling from the endosome, with a clear endosomal colocalisation. Complementation experiments with WT WIPI2 recovered the phenotype, but various mutant WIPI2 complements resulted in elongated tubules, and there was also a dominant negative effect of the mutant. Integrin is a classic retreiver cargo, so the authors rationalise that WIPI2 may be playing a role with retreiver that WIPI1 plays with retromer. To assess this, they perform a set of immunoprecipitations. SNX17, the retreiver-associated sorting nexin, co-IPs with WIPI2 in a VPS26C-dependent manner. VPS26C but not VPS26 co-IPs with WIPI2, and the reciprocal with WIPI1. These interactions were not present for the FSSS mutation of WIPI2. WIPI2 localises to Rab11 endosomes mainly, as does retriever. Mutations of WIPI2 not only affected WIPI2 localisation, but also VPS35L mutations, indicating that there is a functional relationship between the two.

On the whole, I find the manuscript compelling. The manuscript is very clearly written, the results are convincing and well performed. The flow of experiments is logical, and although not comprehensive in the subsequent mechanistic understanding, the fundamental findings are important and convincing. My comments below are, on the whole, minor and are intended to support the communication of the findings to the field.

(1) The IP interaction data were convincing; however, for me and some others, an interaction is only convincing when performed in vitro, and understood at a structural level. I do not suggest the authors do that in this case; however, I think, at a minimum, some sensible moderation of claims would be useful here.

(2) I found the final localisation data and its interpretation confusing. My interpretation of that data would not be that the retreiver is relocalised, but rather that there is less of both recruited to the membrane and the remaining localisation distribution is shifted. In addition, I am not quite sure of the model here - is the idea that WIPI2 recruits retreiver, if that is the case, I find it hard to resolve with its role as a mediator of fission. Clarity would be appreciated here.

(3) I am concerned that the repeats being compared for statistical analysis are not biological repeats but technical repeats (cells in the same experiment). I should think the idea of the statistical comparison is to show experimental reproducibility and variability across biological repeats. Therefore, I would expect an appropriate number of biological repeats (3 or more minimum), to be the data compared in the statistical analysis and graphs. I think it is appropriate to average the technical repeats from each biological repeat. I find these to be useful resources https://doi.org/10.1083/jcb.202401074, https://doi.org/10.1083/jcb.200611141

Reviewer #2 (Public review):

Summary:

The manuscript from De Leo and Mayer presents evidence that the PROPPIN protein, WIPI2, associates with the Retriever complex, and is required for the proper transport of the SNX17-Retriever cargo, beta1-integrin. This finding fits with prior papers from the Mayer lab, which showed that a related PROPPIN, WIPI1, is required for the transport of some SNX27-Retromer cargo, including GLUT1. The retromer and retriever complexes are architecturally similar. Importantly, they act at the same endosomes, and each transports cargo from endosomes to the plasma membrane. Thus, the possibility that each also requires a structurally related PROPPIN is of interest. However, the manuscript is incomplete, and the main claims are only partially supported.

Strengths:

The topic that PROPPIN proteins are important for the function of the Retromer and Retriever complexes expands our view of the trafficking complex.

Weaknesses:

Many important controls are missing. Several points that are made in the manuscript are only supported through a single approach.

Reviewer #3 (Public review):

Summary:

The manuscript of Mayer and colleagues analyzes the function of WIPI proteins in mammalian cells. The authors previously identified CROP as a complex consisting of WIPI1 and the retromer complex, primarily in yeast cells. In mammalian cells, both WIPI1 and WIPI2 exist, whereas retromer has a homologous complex termed retriever. They now find that WIPI2 can form a complex with retriever subunits. They named this complex CROP2. Their data further indicate that CROP2 and CROP1 have distinct substrate specificities as knockdown of CROP2 subunits affects beta1 integrin sorting, whereas knockdown of CROP1 affects EGFR and GLUT1. They further identify a similar sequence (FSSS) in both WIPI1 and WIPI2, which is required for their specific binding to retromer and retriever.

Strengths:

CROP1 and CROP2 seem to use similar features for their formation, and have different substrates, which is convincingly shown.

Weaknesses:

The analysis lacks information that this is a complex as claimed. It can be deduced from the interaction analysis, but was not shown.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

WIPI1 is a PROPPIN family protein that has been implicated in Retromer-mediated membrane fission events. Although the cargos that it has been tested to be important for are diverse, one of the cargos that is unaffected is Beta1-Integrin. This leads the authors to assess another PROPPIN family protein - WIPI2, which is a homolog of WIPI1. KD using siRNA is effective and had no consequences on LAMP1, EGFR trafficking or GLUT1 trafficking. Integrin-B1, however, had a large and significant defect in its recycling from the endosome, with a clear endosomal colocalisation. Complementation experiments with WT WIPI2 recovered the phenotype, but various mutant WIPI2 complements resulted in elongated tubules, and there was also a dominant negative effect of the mutant. Integrin is a classic retreiver cargo, so the authors rationalise that WIPI2 may be playing a role with retreiver that WIPI1 plays with retromer. To assess this, they perform a set of immunoprecipitations. SNX17, the retreiver-associated sorting nexin, co-IPs with WIPI2 in a VPS26C-dependent manner. VPS26C but not VPS26 co-IPs with WIPI2, and the reciprocal with WIPI1. These interactions were not present for the FSSS mutation of WIPI2. WIPI2 localises to Rab11 endosomes mainly, as does retriever. Mutations of WIPI2 not only affected WIPI2 localisation, but also VPS35L mutations, indicating that there is a functional relationship between the two.

On the whole, I find the manuscript compelling. The manuscript is very clearly written, the results are convincing and well performed. The flow of experiments is logical, and although not comprehensive in the subsequent mechanistic understanding, the fundamental findings are important and convincing. My comments below are, on the whole, minor and are intended to support the communication of the findings to the field.

We are happy that the reviewer has received our work quite positively.

(1) The IP interaction data were convincing; however, for me and some others, an interaction is only convincing when performed in vitro, and understood at a structural level. I do not suggest the authors do that in this case; however, I think, at a minimum, some sensible moderation of claims would be useful here.

Indeed, quantitative in vitro data on the affinities would be a nice addition. However, we have significant trouble to recombinantly express and purify well-behaved WIPI2 in sufficient quantities for such studies. We keep working in this direction but are not there yet.

We have now inserted a phrase into the discussion section highlighting this limitation: "Our immunoprecipitation assays cannot distinguish and more detailed structural and interaction studies with pure compounds will be necessary to elucidate the nature of this interaction". We nevertheless think that the the isoform specificity of the IPs, the effect of the point mutations in WIPI2 on these interactions, and the functional effects in vivo lend signficant support to the notion of a complex even if there is no proof of direct binding of WIPI2 to Retriever.

(2) I found the final localisation data and its interpretation confusing. My interpretation of that data would not be that the retreiver is relocalised, but rather that there is less of both recruited to the membrane and the remaining localisation distribution is shifted. In addition, I am not quite sure of the model here - is the idea that WIPI2 recruits retreiver, if that is the case, I find it hard to resolve with its role as a mediator of fission. Clarity would be appreciated here.

We are not quite sure what "final" localisation data the reviewer refers to, but we guess it is Fig. 9. This figure primarily provides in vivo evidence supporting the connection between Retriever and WIPI2. It does this by showing that the S67 substitution shifts both proteins. In WIPI2 wildtype cells, WIPI2 and VPS35L strongly colocalize in Rab11 compartments. S67 substitutions in WIPI2 abolish this localisation; WIPI2 shifts mainly to Rab5 compartments, where VPS35L shows only a moderate increase, and to Rab7 compartments, where VPS35L shows no increase at all.

We do not understand the reviewer's interpretation that less Retriever would be recruited to the membranes in the S67 variants. VPS35L remains completely associated with punctate, presumably membrane-bounded structures also in the mutants, providing no evidence for a detachment from the membrane. The same is observed in a WIPI2 knockdown. Therefore, we did not claim that WIPI2 is the main factor recruiting Retriever to the membrane, for which our experiments yield no hints. This does not exclude that the interaction of WIPI2 could strengthen membrane recruitment, or that two pools of Retriever exist, one interacting with Snx17 and another interacting with WIPI2, and that both link to each other in a coat. We did not dwell on this in the discussion because our experiments cannot distinguish these possibilities and were not conceived to analyse membrane recruitment of Retriever.

(3) I am concerned that the repeats being compared for statistical analysis are not biological repeats but technical repeats (cells in the same experiment). I should think the idea of the statistical comparison is to show experimental reproducibility and variability across biological repeats. Therefore, I would expect an appropriate number of biological repeats (3 or more minimum), to be the data compared in the statistical analysis and graphs. I think it is appropriate to average the technical repeats from each biological repeat. I find these to be useful resources https://doi.org/10.1083/jcb.202401074, https://doi.org/10.1083/jcb.200611141

The repeats being compared are biological repeats from independent experiments. This is described in Methods, where the reviewer may not have seen it. In order to make the independent experiments more evident in the figures, we have now colour coded the individual cell measurements from the three independent experiments. This allows to visualize both the individual data points, the average from each experiment and the variability across the independent experiments.

Reviewer #2 (Public review):

Summary:

The manuscript from De Leo and Mayer presents evidence that the PROPPIN protein, WIPI2, associates with the Retriever complex, and is required for the proper transport of the SNX17-Retriever cargo, beta1-integrin. This finding fits with prior papers from the Mayer lab, which showed that a related PROPPIN, WIPI1, is required for the transport of some SNX27-Retromer cargo, including GLUT1. The retromer and retriever complexes are architecturally similar. Importantly, they act at the same endosomes, and each transports cargo from endosomes to the plasma membrane. Thus, the possibility that each also requires a structurally related PROPPIN is of interest. However, the manuscript is incomplete, and the main claims are only partially supported.

Strengths:

The topic that PROPPIN proteins are important for the function of the Retromer and Retriever complexes expands our view of the trafficking complex.

Weaknesses:

Many important controls are missing. Several points that are made in the manuscript are only supported through a single approach.

We made a serious effort and implemented many suggestions of this reviewer, but orthogonal approaches are not always available or accessible.

Reviewer #3 (Public review):

Summary:

The manuscript of Mayer and colleagues analyzes the function of WIPI proteins in mammalian cells. The authors previously identified CROP as a complex consisting of WIPI1 and the retromer complex, primarily in yeast cells. In mammalian cells, both WIPI1 and WIPI2 exist, whereas retromer has a homologous complex termed retriever. They now find that WIPI2 can form a complex with retriever subunits. They named this complex CROP2. Their data further indicate that CROP2 and CROP1 have distinct substrate specificities as knockdown of CROP2 subunits affects beta1 integrin sorting, whereas knockdown of CROP1 affects EGFR and GLUT1. They further identify a similar sequence (FSSS) in both WIPI1 and WIPI2, which is required for their specific binding to retromer and retriever.

Strengths:

CROP1 and CROP2 seem to use similar features for their formation, and have different substrates, which is convincingly shown.

Weaknesses:

The analysis lacks information that this is a complex as claimed. It can be deduced from the interaction analysis, but was not shown.

It is of course desirable to obtain a detailed structural and in vitro characterisation of this interaction, which we have not provided because we currently do not have sufficient amounts of well-behaved source material for this. We nevertheless think that the interaction we show, which is strictly isoform-specific and dependent on single amino acid substitutions in a motif that in CROP1 is necessary for the interaction its recombinant subunits, supports that CROP2 is a similar a complex. We don't show a direct interaction but also don't claim in the manuscript that the interaction between WIPI2 and Retriever is direct and independent of additional factors.

Recommendations for the authors:

Reviewing Editor Comments:

As you will see, the reviewers generally value the contribution to the field, but they feel that some claims require additional experimental support.

(1) I have summarized the major points below.

(a) Both reviewers 1 and 2 agree that the quality of localization data presented in Figure 9 and S5-S7, and the interpretation of the data, could be improved. See comment 2 from reviewer 1 and comments 23, 24 and 25 from reviewer 2. They not only suggest ways to improve the presentation of the data, but additionally suggest improving the staining of the Rab11 marker and additionally explain the lack of co-localization between VPS35 and Rab5, which has been reported in the literature.

This impression was due to the fact that some figures showed projections of image stacks, which was not indicated clearly in the figure legend. We have changed this and now show single image planes throughout all figures.

(b) Both reviewers 1 and 3 note that the evidence supporting a functional WIPI2-Retriever complex in vivo is currently weak. We agree that additional biochemical data demonstrating the presence of the CROP1 and CROP2 complexes in vivo would strengthen the central message of the paper and elevate it to a more fundamental discovery.

We understood that the reviewers did not ask for further in vivo evidence but would welcome structural characterisation of the complex and quantitative binding data in vitro with purified proteins. Structural characterisation is out of scope of our study and in vitro binding studies have remained hampered by the fact that WIPI2 is hard to express and purify and not well behaved in vitro.

(c) All reviewers agree that the authors should carefully repeat their statistical analysis to account for the number of biological replicates. Reviewer 1 suggests publications that the authors could refer to.

The reviewers have probably overlooked the respective description in the methods section, where it had been stated that we analysed biological replicates from independent experiments. In graphs showing measurements from individual cells we now make this evident through colour coded dots, in which each colour represents data points stemming from an independent experiment. This makes it evident that the variance from experiment to experiment is low. The means (n = 3) were generally compared using a two-tailed unpaired t-test.

(d) Reviewer 2 additionally has various minor points that would greatly improve the readability and presentation of the work, and we recommend addressing (comments 1, 2, 3, 4, 12, 15, 17, 20, 27, 28, 29). All reviewers, in general, provide great minor suggestions. It would be great if the CROP1 and 2 complexes could be clearly introduced in each figure. We also agree that the WIPI2 CT labelling is confused and should be changed to "control" or similar.

Many of the points raised by this reviewer were actually quite minor or questions of personal preference, not major problems as stated in the review. Nevertheless, we found a number of useful suggestions in this review and have addressed these points as detailed in the response to reviewer 2.

(2) In addition to the major shared concerns laid out in the points above, reviewer 2 has some further minor suggestions:

(a) Comment 6. Could the author explain the discrepancies between the example blot shown in Figure 1D and the quantification (1E).

The two have actually been quite consistent. The reviewer might have mistaken the marker lane as the 0 min reference value to arrive at this impression. We have now removed the marker lane to avoid this.

(b) Comment 9 - could the authors clarify how surface labelling experiments were carried out?

This had been clearly described in the methods section, where this reviewer has probably not seen it.

(c) Comment 11 - The reviewer suggests normalizing the surface levels of markers to the cell area and not per cell. This is a reasonable suggestion.

The analysis had already been performed as proposed. This had been clearly described in the methods section, which the reviewer may not have looked at.

(d) Comment 19 "In Figure S4, the authors observe tubular structures. The authors should perform immunofluorescence with endosomal markers such as EEA1, LAMP1 and Retromer to determine the nature of the tubulovesicular structures." The authors could try a Rab4 or Rab11 overexpression plasmid to show whether these are elongated recycling tubules.

This has now been added.

Reviewer #1 (Recommendations for the authors):

Minor comments:

(1) The figures are not colourblind friendly, and should be changed to be so. Additionally, single colour images should be grayscale.

That was a good learning opportunity. We adapted the colour schemes of the images to make them more colourblind friendly, now using magenta, green, and white for the overlaps. In doing so we have relied on published recommendations, but we have not found a colourblind colleague to check the efficacy of this change.

(2) WIPI2^CT labels are confusing, as people may think they are a mutant. I suggest changing to "control" or similar.

These have been changed.

(3) "The effect was comparable to that of a knockdown of SNX17 (Figure 3 A, B)." On page 6. Based on this sentence, I was expecting to see a comparison to SNX17 KD, but it was not there as far as I can tell.

This statement referred to a publication by P.Cullen and collaborators. We have changed the wording and inserted the (missing) reference to make this clear.

Reviewer #2 (Recommendations for the authors):

The manuscript is modest. In addition, many of the claims should be better supported by the addition of orthogonal data. Moreover, the quality of some of the data presented needs to be improved. Overall, the manuscript requires better descriptions of the methods. In many figures, it was not clear how the experiments were performed.

The experimental descriptions that the reviewer refers to had been provided in the Methods section, where this reviewer may have overlooked them.

The paper should also be better organized. Some less important findings are in the main figures, whereas some critical results are in the supplemental figures. In addition, there were multiple issues with the readability of the paper, and the authors should consider using a professional editor to make the paper easier to read.

We had given the paper to colleagues who found it clear, and also Reviewer 1 has underlined its clarity. Nevertheless, we have re-phrased the manuscript in some parts to optimise it.

One of the main claims in the paper is that the FSSS motif of WIPI2, as well as a conserved amphipathic helix, is critical for WIPI2 function in the CROP2 complex. It is notable that these are the same regions that are also critical for the role of WIPI2 in autophagy (Gubas et al., 2024 PMID: 39152217). The authors should include this information in the manuscript and cite the paper.

Indeed. We mention this now in the introduction of the revised version.

Additional Major Issues:

While some of the issues raised below are actually minor and/or matters of personal preference, several comments led us to improve and correct the figures and we thank this reviewer for the constructive suggestions.

(1) In Figure 1, it appears from the representative images that WIPI2 KD cells have higher levels of EGFR (Figure 1A and 1B). Is this correct?

To some degree. This increase is not systematic. A moderate increase has been observed only in 2 experiments out of 4. Therefore, we did not investigate this.

(2) Also in Figure 1, the colocalization is difficult to see. The authors should add the separate channels in addition to the merged images. Since the point is supposed to be that there is no impact on EGFR, all of this data could go into the supplement.

We had considered this already for the original version but dismissed the idea. The overlap is quantified in Fig. 1C, which provides the relevant values from four experiments. Fig. 1A/B provide only sample pictures, which also permit to see overlap (yellow) 0 and 5 min after the induction of degradation, which vanishes at later timepoints. Separating the channels would quadruple the space that this figure occupies, which would not be practical and not change the point to be made.

(3) The scale bars for each panel differ from each other. To better assess the data, the exact same magnification should be shown for each panel.

Corrected

(4) Figure 1C is confusing. The authors should explain which lines correspond to EEA1 and LAMP1.

Corrected

(5) In Figure 1D, the authors show different blots for control and WIPI2 KD. Could the authors compare WIPI2 and EGFR in the same blot? Without a comparison on the same blot, it is impossible to know whether the starting levels of EGFR are the same. Moreover, the quantitation in Figure 1E sets the value for each cell line to 100%. Instead, the starting levels in each cell line should be compared. The authors should use the amount of EGFR at zero time in the control cells to define 100%, and then indicate the relative initial EGFR levels in the WIPI2KD cells.

A new blot is shown now and the quantification has been performed as proposed.

(6) The quantification in Figure 1E does not match the representative blot shown in Figure 1D. According to the graph, the rate of degradation of EGFR is similar in both cell lines. But the representative blot shows that there are large differences.

We do not understand this comment. The representative blot shows similar kinetics for both. Perhaps the reviewer got confused by the fact that a marker lane was still present on the left blot and not labelled as such. The new version of the figure corrects this.

(7) The blot showing the WIP2 knockdown in Figure 1D has a lot of background. However, the blot of the WIPI2 knockdown in Figure S1 looks very good. The authors should make sure that they load enough sample and use a good antibody for the experiments in Figure 1.

The new blot that we added in response to comment 5 corrects this.

(8) In Figure 2 and Figure 3A, the cells are too confluent. This is an issue because the cells might not be metabolically active. In addition, the signal is saturated. The authors should make sure that all of the data is collected on cells that are not too confluent.

The confluency of the culture cannot be judged from single frames, which were selected to show several cells. We had controlled confluency and underlined in the Methods section that “For microscopy, the cells were plated on 18-mm-diameter glass coverslips on 24-well plates and grown for 2 or 3 days according to the protocol of DNA or siRNA transfection by reaching a confluency of 70-80%”. The reviewer may not have seen this.

(9) One main issue with these figures, especially the non-permeablized cells, is that it is impossible to assess how much of the signal is on the cell surface. The authors should provide the methods that they used to prevent inadvertent permeabilization of the cells. Were these experiments performed at 4 degrees? The authors should include a control of an antibody to a protein that is not found on the cell surface.

There is an internal control in that the non-permeabilised WIPI2KD cells, which have been treated with the same antibody, show no much less staining than the control cells (Fig. 3A). In WIPI2KD cells, integrin becomes accessible for antibody staining only upon detergent permeabilization. This demonstrates that our procedure does not lead to significant inadvertent permeabilization of the cells.

(10) The authors should perform surface biotinylation assays as an orthogonal approach to determine GLUT1 levels and beta1-integrin levels at the cell surface, respectively.

There is a strong, qualitative difference in the surface labelling of beta1-integrin that is not observed for GLUT1. Given that, it is not obvious to us what additional argument would be provided by surface biotinylation or subfractionation experiments.

(11) In quantifying surface levels of GLUT1 or beta1-integrin by microscopy, the authors should normalize to the cell area, rather than per cell.

The reviewer has probably not seen that the Methods section states that the cell area has been used for normalisation.

(12) In Figure 3, the nuclear DAPI stain in the KD cells is much less bright than in the control cells. The authors should make sure to choose representative images.

The nuclear DAPI signal has been visible in all cells. Depending on the position of the nucleus, is shape and dimension in the z-direction, individual nuclei can show different degrees of staining. The images shown are representative. We have adjusted the settings now to make the nuclei in the WIPI2KD cells easier to spot.

(13) For the immunofluorescence studies, the authors should be using single z planes rather than maximum projection.

Images have been exchanged by single planes.

(14) For the experiments in Figure 3, the authors should check the total levels of EEA1 and LAMP1 by western blot to test whether WIPI2 KD affects the levels of these proteins. If these organelle marker proteins are impacted, this could impact the colocalization measurements shown in Figures 3C and D.

We have measured the total fluorescence intensity of EEA1 and LAMP1 in the images. It shows no significant difference between control and WIPI2 knockdown cells (new Fig. 3F, H).

(15) In Figure 4A, the helical representation is rotated in the WIPI2-Sloop; the orientation of the residues that are not mutated should stay the same.

Yes. Done.

(16) In Figure 4B and 4C, cells that were not transfected with WIPI2 WT or WIPI2 Sloop should be shown.

Since the transfection efficiency is limited, the fields contain both non-transfected (lacking green fluorescence) and transfected cells (showing green fluorescence). We have now marked transfected cells with an asterisk.

(17) The cells in the lower panel of 4B have an unusual morphology and are much more round. The authors should choose cells that are representative of each experimental condition.

We now provide another field.

(18) In Figure 4C, it looks like the magnification of the top panels is different from the bottom panels. The same magnification for all the panels should be shown (and the size of the scale bars should be the same.

Corrected

(19) In Figure S4, the authors observe tubular structures. The authors should perform immunofluorescence with endosomal markers such as EEA1, LAMP1 and Retromer to determine the nature of the tubulovesicular structures.

We have done this (new Fig. S4). Rab4 is on tubules. Rab5 on the structures from which the tubules emanate.

(20) In Figure 5A, the top scale bar is missing.

Corrected.

(21) In Figure 5B, the confluency is too high.

See our response above. A single field does not permit to judge this. Confluency was controlled for all cultures. The cultures were not confluent.

(22) The IP studies shown in Figures 6, 7 and 8, should be accompanied by colocalization studies.

Colocalization measurments have now been integrated into the manuscript (Figs. S5, S6). They are consistent with the IP data.

(23) Figure 9 was very confusing and should be broken up into multiple figures. Data showing that localization did not change in any of the cell lines can be put in figures that are distinct from figures that show that localization changed in the various mutants. Figures that show no change can go in the supplement.

Since every panel of Fig. 9 shows a statistically significant difference we left the figure unchanged.

(23) Representative figures should be shown in the same figure as the corresponding graph. In addition, the order of the colocalization data shown in the graphs and figures should match the order described in the text.

We consider the graphs of Fig. 9 as the relevant information. Representative images are just illustration. Integrating them with the graphs would make it necessary to split everything up into multiple figures, making it harder to compare the different combinations. Therefore, we left the figures unchanged.

(24) In Figure S7, the Rab11 signal looks continuous, which makes the colocalization analysis meaningless. The authors should determine how to take images that can be evaluated. On a more minor note, the zoomed panels should be labeled as well.

This is a result of having shown a projections of multiple planes. The images have now been replaced by single plane images. Zoomed panels have been labelled and the scale bar added.

(25) The low colocalization of VPS35L with Rab5 is surprising, as SNX17 has been previously shown to co-localize with early endosomes positive for EEA1. This result may have occurred due to overexpression because the authors chose to utilize plasmids that express a tagged protein. There are antibodies to each of the endogenous proteins, and this is what should be used for this set of experiments.

This comment made us control the analysis performed for these images, which by mistake had been performed on z-projections rather than on single planes. This distorted the values. The re-analysed data shows a higher colocalisation with Rab5, but it remains inferior to colocalisation with Rab11.

(26) The authors should determine whether β1-integrin colocalizes with WIPI2 in endosomal compartments.

This was done. WIPI2 colocalizes with beta-integrin on EEA1-and SNX17-positive strcutures but not positive for LAMP1 (Fig. 3E/F).

Minor points

(27) In one of the panels in Figure 1A, "30 min" is duplicated.

Removed

(28) In Figures 5C and 5D, the y-axis should indicate that this is surface β1integrin.

Changed and added “surface”

(29) In Figure 9 there is a typo in panel A. It is VPS35L and not VPS35.

Corrected

Reviewer #3 (Recommendations for the authors):

This is an overall convincing study, which shows that the two complexes, CROP1 and CROP2 function at different membranes and serve different substrates. While I agree with their localization analysis, I have one key issue. The authors claim that each of the two forms a complex and base this on their specific pull-down and western blot analyses.

I find it important that they show that both indeed form stable complexes in vivo, using pull-down and mass spectrometry approaches. They have all the necessary tools in hand and could use WIPI1 and WIPI2 to demonstrate the existence of the two complexes. The FSSS mutants of each are good controls for such an analysis.

The manuscript actually presents the demanded in vivo experiments. Figs. 6 to 8 show pull-downs of WIPI1 and WIPI2 from cells, including also the FSSS mutant. While we haven't analysed this interaction by mass spectrometry, the Western blot analysis confirms the analysis. Cooperation of these proteins is further supported by the in vivo phenotypes, where the S67A substitution in WIPI2 produces a similar phenotype on integrin beta1 localisation as inactivation of Retriever.

A second aspect is the general presentation. The paper would be a lot more accessible if the subunits of each complex (CROP1 and CROP2) were also introduced in the figures of each part. For readers, a final model is helpful to put the data into context and show where each complex operates in the cell.

We have introduced a scheme of the respective complexes, including the names of the compunds, in Figs. 6 and 7 to avoid confusion.

Finally, it is not clear how the statistics compare to repeats in their data. This should be clarified.

This had been described in methods. Statistics has always been done on biological replicates stemming from independent experiments. We have added a cartoon (Fig. 10) depicting the trafficking pathways affected by CROP1 and CROP2.

Version 2

Abstract

Endosomes generate tubulo-vesicular carriers to redistribute proteins between plasma membrane, Golgi, and lysosomes. These transport routes employ distinct combinations of sorting nexins with complexes such as Retromer or Retriever. We now show that, while Retromer associates with the PROPPIN WIPI1 to form the previously described CROP complex, Retriever associates with WIPI2, forming CROP2. WIPI2 integrates into Retriever-dependent coat complexes, since it interacts both with the Commander subunit CCDC93 and its cognate sorting nexin SNX17. CROP and CROP2 are exclusive in their physical associations and pathway selective. Whereas CROP2 is required for endosomal exit of β1-Integrin, it does not affect CROP-dependent cargos, such as EGFR or GLUT1. Vice versa, CROP is not required for β1-Integrin trafficking. WIPI1 and WIPI2 rely on similar molecular features. Their activity depends on the same FSSS motif to integrate into Retromer and Retriever complexes, respectively, and on an amphipathic membrane-inserting α-helix, which conveys membrane fission activity to PROPPINs. This suggests that Retromer and Retriever coats integrate distinct PROPPIN isoforms to promote fission of the respective endosomal carriers formed by them.

Introduction

Endosomal membrane traffic regulates the abundance and activity of many receptors and transporters at the plasma membrane, which renders it critical for the development and homeostasis of cells and tissues (McNally and Cullen, 2018). In line with this, perturbations in endosomal trafficking contribute to a wide range of diseases, including cancer and some of the most common neurological disorders (Carosi et al., 2023; Mellman and Yarden, 2013; Schreij et al., 2016). Many plasma membrane proteins cycle between the cell envelope, endosomes, and the Golgi. They can move on to lysosomes for degradation unless they are retrieved before endosomes fuse with lysosomes (Cullen and Steinberg, 2018). Endosomes thus serve as sorting stations at an intersection between several compartments. To collect cargo proteins, endosomes employ a variety of coats that form tubulo-vesicular carriers. Many of them are based on sorting nexins, which scaffold the membrane into narrow tubules. Sorting nexins can act on their own, such as in the ESCPE-1 complex (Kvainickas et al., 2017; Lopez-Robles et al., 2023; Simonetti et al., 2019, 2017), or together with complexes interconnecting them, such as Retromer or Retriever and the Commander complex, into which Retriever can be integrated (Bartuzi et al., 2016; Boesch et al., 2023a; Chandra et al., 2024; Healy et al., 2023; Mallam and Marcotte, 2017; McNally et al., 2017; Phillips-Krawczak et al., 2015; Seaman, 2021). Further coats, such as clathrin adapter complexes, VINE, and ACAP1-clathrin can also act on endosomes (Dai et al., 2004; Hooy et al., 2022; Li et al., 2007; Pang et al., 2014; Ren et al., 2013; Sachse et al., 2002; Shortill et al., 2022). These coat components organize a variety of endosomal protein exit routes with partially overlapping cargo specificity (Bean et al., 2017; Kvainickas et al., 2017; McGough et al., 2014; Steinberg et al., 2013), which may reflect their capacity to form hybrid carriers integrating different classes of sorting nexins (Gopaldass et al., 2026) .

Retromer is part of a protein coat mediating protein exit from endosomes in association with sorting nexins, such as Vps5, Vps17 or Snx3/Grd19 in yeast, and SNX27, ESCPE-1 or SNX3 in human cells (Seaman, 2021). Retromer consists of a VPS26–VPS29–VPS35 heterotrimer, which can form higher order complexes through VPS35 and VPS26 dimerization (Collins et al., 2008; Deatherage et al., 2020; Hierro et al., 2007; Kendall et al., 2022, 2020; Kovtun et al., 2018; Leneva et al., 2021; Lucas et al., 2016; Purushothaman et al., 2017). While sorting nexins themselves change the spontaneous curvature of membranes and can tubulate them (Kovtun et al., 2018; Leneva et al., 2021; Weering et al., 2012; Zhang et al., 2021), this activity can be enhanced by the association with Retromer, which may provide additional driving force for scaffolding the membrane (Gopaldass et al., 2023).

In human cells, numerous cargos recycle back to the cell surface in a Retromer-independent manner (Steinberg et al., 2013). Many of these depend on Retriever, an evolutionarily conserved complex of VPS26C/DSCR3, VPS29 and VPS35L/C16orf62, proteins that show distant homology to the Retromer subunits (McNally et al., 2017). Whereas the overall organization and shape of Retriever resemble those of Retromer, Retriever shows different cargo specificity and distinct features. Retriever forms part of a much larger complex, Commander, which integrates Retriever with the 12-subunit CCC complex and DENND10 and cooperates with the sorting nexin SNX17 to form coats (Boesch et al., 2023a; Butkovič et al., 2024; Healy et al., 2023; Martín-González et al., 2024; Singla et al., 2024; Steinberg et al., 2012). Like Retromer, also Commander can interact with WASH and thereby link to actin (Derivery et al., 2009; Gomez and Billadeau, 2009; Guo et al., 2024; Harbour et al., 2012; Jia et al., 2012; Phillips-Krawczak et al., 2015; Seaman et al., 2009).

Cargo carriers detach from endosomes by membrane fission (Gopaldass et al., 2024; Naslavsky and Caplan, 2023). Several factors have been implicated in this detachment, among them actin and the actin-controlling WASH complex (Derivery et al., 2009; Frisby et al., 2024; Gomez and Billadeau, 2009; Harbour et al., 2012; Jia et al., 2012; Seaman et al., 2009), dynamin-like GTPases and ATPases such as Vps1 and EHD1 (Arlt et al., 2015; Chi et al., 2014; Daumke et al., 2007; Deo et al., 2018; Gokool et al., 2007; Grant et al., 2001; Solinger et al., 2020), the ESCRT-III subunit IST1 (Clippinger et al., 2024) and the PROPPIN (β-propeller that binds phosphoinositides) WIPI1 and its yeast homolog Atg18 (DeLeo et al., 2021; Gopaldass et al., 2024, 2017). Coats and actin can exert forces on the underlying lipid bilayer, which not only allow the tubule to grow but may also produce friction between lipids and protein coats that contributes to fission (Johannes et al., 2014; Simunovic et al., 2018, 2017). In addition, dynamin-like proteins actively deform the bilayer to compress and destabilize it (Antonny et al., 2016). The lipid phosphatidylinositol-(3,5)-bisphosphate (PI(3,5)P2) and/or PI5P is also required for fission (DeLeo et al., 2021; Giridharan et al., 2022; Gopaldass et al., 2017; Rivero-Ríos and Weisman, 2022; Zieger and Mayer, 2012). PROPPINs are bona-fide effectors of PI(3,5)P2 (Courtellemont et al., 2022; DeLeo et al., 2021). They associate with membranes through two phosphoinositide binding sites (Baskaran et al., 2012; Krick et al., 2012; Liang et al., 2019; Proikas-Cezanne et al., 2004; Vicinanza et al., 2015; Watanabe et al., 2012). Upon membrane binding, a hydrophobic loop, which is located between the two lipid binding sites, folds into an amphipathic α-helix that inserts into the bilayer. Thereby, PROPPINs can tubulate lipid bilayers and drive their fission (Courtellemont et al., 2022; DeLeo et al., 2021; Gopaldass et al., 2017; Gopaldass and Mayer, 2024).

PROPPINs are a protein family that binds phosphatidylinositol 3-phosphate (PI3P), phosphatidylinositol 5-phosphate (PI5P) and phosphatidylinositol 3,5-bisphosphate (PI(3,5)P2) (Baskaran et al., 2012; Busse et al., 2015; Dove et al., 2004; Gopaldass and Mayer, 2024; Jeffries et al., 2004; Vicinanza et al., 2015). They are conserved from yeast (Atg18, Atg21, and Hsv2) to humans (WIPI1, WIPI2, WIPI3/WDR45B and WIPI4/WDR45) and exist in most eukaryotes in several isoforms, which can be grouped into two evolutionary clades (Dove et al., 2004; Proikas-Cezanne et al., 2015). In mammals, Atg18, Atg21, WIPI1 and WIPI2 belong to one clade, Hsv2, WIPI3 and WIPI4 to the second (Cong et al., 2021; Polson et al., 2010; Strong et al., 2021). WIPI1 and WIPI2 concentrate on phagophores when autophagy is induced and they are necessary for efficient formation of autophagosomes (Dooley et al., 2014; Polson et al., 2010; Proikas-Cezanne et al., 2004). WIPI2, which has a stronger impact on autophagy, interacts with different key factors for autophagosome formation.

WIPI1 and Atg18 also support membrane fission and protein sorting on endosomes and lysosomes, but for these functions they rely on molecular features that are distinct from those important to autophagy (Courtellemont et al., 2022; DeLeo et al., 2021; Gopaldass et al., 2017; Gopaldass and Mayer, 2024). WIPI1 localizes to endo-lysosomal compartments, where it affects the size of endosomes and the exit of proteins from endosomes to lysosomes (EGF-receptor), to the Golgi (Shiga-toxin), or to the cell surface (Transferrin receptor, GLUT-1) (DeLeo et al., 2021; Jeffries et al., 2004). Based on the differential effects of substitutions in the two phosphoinositide binding sites of WIPI1 it was suggested that binding of PI3P allows WIPI1 to support the formation of endosomal transport carriers while PI(3,5)P2 and lipid binding site two are needed for membrane fission at endosomes (DeLeo et al., 2021). WIPI1 as well as its yeast homolog Atg18 associate with Retromer to form the CROP complex, which amplifies the fission activity of the PROPPIN and makes its binding to membrane PI(3,5)P2 dependent (Courtellemont et al., 2022; Marquardt et al., 2023). This interaction depends on a conserved FSSS motif, which can be phosphorylated (Courtellemont et al., 2022; DeLeo et al., 2021; Feng et al., 2015; Gubas et al., 2024; Strong et al., 2021). This phosphorylatable FSSS motif exists also in WIPI2, where it forms also part of the binding site for ATG16L1, a core component of the autophagic machinery (Gubas et al., 2024; Strong et al., 2021) .

We have recently identified the Retromer- and WIPI1-containing CROP complex as an agent supporting the export of multiple cargos from endosomes, comprising both Retromer-dependent and Retromer-independent exit routes (Courtellemont et al., 2022; DeLeo et al., 2021). Since not all cargo exit from endosomes depends on WIPI1 and CROP we analysed alternative PROPPINs that might be involved in WIPI1-independent exit from endosomes. This led us to uncover a novel role for a WIPI2-Retriever complex in trafficking of β1-Integrin, which we present below.

Results

WIPI1 promotes the exit of diverse cargos from endosomes, for example for transporting Transferrin to the plasma membrane, GLUT1 to the plasma membrane, EGF-receptor (EGFR) to lysosomes, and of Shiga toxin to the Golgi (DeLeo et al., 2021). β1-Integrin provides an exception, its delivery to the cell surface being unaffected by knockout of WIPI1 (Courtellemont et al., 2022). We investigated whether the endosomal exit of β1-Integrin might depend on another PROPPIN. Since WIPI1 and WIPI2 are similar (Bakula et al., 2013; Gopaldass and Mayer, 2024) belong to the same evolutionary clade, share two predicted lipid binding sites as well as the potential to form an amphipathic helix on their CD loop 6 (Cong et al., 2021; Gopaldass et al., 2017; Polson et al., 2010), we focused our attention on WIPI2B.

WIPI1 and WIPI2 are specific for distinct trafficking pathways

We tested the impact of WIPI2 knockdown (WIPI2KD) on the afore-mentioned trafficking pathways. siRNA against WIPI2 reduced the protein in whole cell extracts by 94% when compared to cells transfected with control siRNA (Control) (Figure. S1). We analysed trafficking of EGFR from early endosomes to degradative, LAMP1-containing compartments. After 24h of serum starvation, cells were stimulated with EGF and analysed for EGFR localization by immunofluorescence. 15 min after EGF addition, EGFR-positive structures colocalized with the early endosomal marker EEA1 to more than 75% in both WIPI2KD and control cells, indicating that EGF had induced efficient internalization of EGFR in both cases (Figure 1A-B). After 30 and 60 min of EGF stimulation, EGFR increasingly shifted from EEA1- towards LAMP1-positive lysosomal structures in control as well as in WIPI2-silenced cells. EGFR signals declined after 60 min in the immunofluorescence images as well as in Western blot analyses of the both cells (Figure. 1 C-E). Thus, WIPI2 knockdown does not impair EGFR trafficking from EEA1 endosomes to LAMP1-positive compartments and leaves lysosomal degradation of EGFR functional.

Effect of WIPI2 knockdown on EGFR degradation.

A. B. Control and WIPI2KD cells were serum-starved for 24 h and then supplemented with EGF (100 ng/ml). After the indicated periods of time, cells were fixed, permeabilized, DAPI-labeled (blue) and decorated with antibodies to EGFR (magenta), EEA1 or LAMP1 (green). Scale bars: 10 μm. C. Colocalization of EGFR with EEA1 or LAMP1 (white) was quantified over time using the images from A and B and Manders’ correlation coefficients were calculated. M1 indicates the fraction of magenta pixels overlapping with green pixels. Values are the mean ± s.d.; (n = 3). 150 cells, stemming from 3 independent biological experiments, were quantified per sample. D. EGFR degradation. Control and WIPI2KD cells were stimulated with EGF for the indicated periods of time, lysed and subjected to SDS-PAGE and Western blot analysis for EGFR. α-Tubulin served as a loading control. E. Quantification of EGFR from (D), using the value of control cells at time 0 (cells starved for 24 h) as 100% reference. Data are means ±s. d., from three independent experiments. Data points form these three experiments are indicated for each time point.

To test the role of WIPI2 in Retromer-dependent trafficking we assayed the localisation of the glucose importer GLUT-1 because this protein accumulates on endosomes if its recycling to the plasma membrane through Retromer is impaired (Steinberg et al., 2013). To capture the surface fluorescence, we acquired z-stacks of non-permeabilized, immuno-stained cells, generated maximum intensity projections and integrated the fluorescence signal over the entire cell, normalized for its area as described in Materials and Methods section. Non-permeabilized WIPI2KD cells showed similar cell surface staining of GLUT-1 as control cells (Figure 2 A-B), suggesting that also Retromer-mediated recycling of GLUT1 occurs independently of WIPI2.

GLUT1 expression and localization upon WIPI2 knockdown.

A. GLUT1 cell surface exposure. Control and WIPI2KD cells were fixed and stained with antibody against GLUT1 and with DAPI. Where indicated, cells had been permeabilized with 0.05% before staining to reduce plasma membrane staining and provide better access to GLUT1 inside the cell. Scale bars: 10 μm. B. Quantification of GLUT1-immunofluorescence in cells from A. Regions of interest (ROIs) corresponding to each cell and in some regions outside the cells (background) were manually defined using image J software. Total cell fluorescence was integrated, and the background fluorescence was subtracted. The resulting total cell fluorescence was divided by the area of the cell. This value is shown in the graph. 150 cells per condition, stemming from three independent experiments, were quantified. Values of individual cells and the means are presented by smaller and larger circles, respectively, coloured according to the independent experiment that generated them. P values were calculated applying an unpaired Student’s t-test with unequal variances. The analysis was performed with 99% confidence. NS = not significant (P > 0.05).

Next, we tested the requirement of WIPI2 for trafficking of β1-Integrin. Surface accumulation of this Integrin subunit is unperturbed by loss of SNX27 or Retromer but depends on the SNX17-Retriever pathway (McNally et al., 2017; Steinberg et al., 2013). Knockdown of WIPI2 induced a loss of β1-Integrin immunofluorescence from the surface of HK2 cells (Figure 3 A, B) and phenocopied the previously published knockdown of SNX17 (McNally et al., 2017; Steinberg et al., 2013). When the cells were detergent-permeabilized to partially remove the plasma membrane and improve accessibility of internal structures for immunostaining, cells depleted of WIPI2 showed an accumulation of β1-Integrin on intracellular structures that colocalized with the early endosomal marker EEA1 and, to a lesser degree, with the late endosomal/lysosomal marker LAMP1 (Figure 3 C-E,G). WIPI2 knockdown had no impact on the total EEA1 and LAMP1 fluorescence level per cell, suggesting that the abundance of these compartments remained the same (Figure 3 F,H). In accord with previous observations (DeLeo et al., 2021), β1-Integrin surface labelling in non-permeabilized cells was not affected by knockdown of WIPI1 (Figure S2). Thus, β1-Integrin recycling to the plasma membrane selectively requires WIPI2.

Impact of WIPI2 knockdown on β1-Integrin localization.

A. β1-Integrin cell surface expression. Control and WIPI2KD cells were fixed and stained with antibody to β1-Integrin (green) and with DAPI (blue), without detergent permeabilization. Scale bars: 10 μm. B. Quantification of β1-Integrin immunofluorescence in cells from A. Regions of interest (ROIs) corresponding to each cell and to areas outside the cells (background) were manually defined using ImageJ software. Total cell fluorescence was integrated, and the background fluorescence was subtracted. The resulting total cell fluorescence was divided by the area of the cell. This value is shown in the graph. 150 cells per condition, stemming from three independent biological experiments, were quantified. Individual cells and means are presented by smaller and larger circles, respectively, coloured according to the independent experiment that they stem from. P values were calculated through an unpaired Student’s t-test with unequal variances.The analysis was performed with 99% confidence. ***P < 0.001. C.D. Immunofluorescence staining of intracellular β1-Integrin (green) and EEA1 (magenta) or LAMP1 (magenta) in control and WIPI2KD cells. Overlaps are marked in white. Cells had been fixed, permeabilized with 0.05% saponin and stained with antibodies to the indicated proteins. Scale bars: 10 μm. E.G. Colocalization of β1-Integrin with LAMP1 or EEA1 was measured in cells from C, using Manders’ colocalization coefficient M2, calculated in Image J software. It indicates the fraction of green pixels overlapping with the magenta pixels. Colocalization was quantified from three independent experiments with a total of 150 cells per condition. Values of individual cells and means are presented by smaller and larger circles, respectively, coloured according to the independent experiment that they stem from. An unpaired Student’s t-test with unequal variances was used to calculate P-values. The analysis was performed with 99% confidence. ***P < 0.001. F.H. Quantification of LAMP1- or EEA1-immunofluorescence in cells from C or D, respectively, was performed as in B. 120 cells per condition, stemming from three independent biological experiments, were quantified. Individual cells and means are presented by smaller and larger circles, respectively, coloured according to the independent experiment that they stem from. Data are means ± s.d. P-values were calculated applying an unpaired Student’s t-test with unequal variances. The analysis was performed with 99% confidence. NS = not significant (P > 0.05).

The conserved amphipathic α-helical and FSSS motifs of WIPI2 promote β1-Integrin recycling

WIPI2 shares two critical features with other PROPPINs (Gopaldass and Mayer, 2024). These include an FSSS motif and the CD loop on blade 6 of the β-propeller structure. The CD loop promotes membrane fission activity of Atg18 during the fragmentation of yeast vacuoles (Gopaldass et al., 2017) and in the detachment of endosomal carriers (DeLeo et al., 2021). Upon contact with the bilayer, the loop folds into an amphipathic α-helix that is conserved among PROPPINs and necessary for membrane fission activity. To test whether the amphipathic nature of this helix also contributes to the function of WIPI2 in β1-Integrin retrieval to the cell surface, we swapped two pairs of amino acids on opposite sides of this α-helix in WIPI2 (Figure 4 A). This manipulation preserves the amino acid composition of the helix but abolishes its amphipathic character. The resulting EGFPWIPI2SLoop (scrambled loop) variant and the wildtype version EGFPWIPI2WT were expressed in control and WIPI2KD cells at comparable levels, as shown by Western blot (Figure S3). Whereas permeabilized WIPI2KD cells displayed β1-Integrin concentrated on perinuclear endosomes, expression of EGFPWIPI2WT rescued this phenotype (Figure 4 B). EGFPWIPI2Sloop failed to induce such recovery in WIPI2KD cells. It even exerted a dominant-negative effect when expressed in control cells. (Figure 4 C-D). Cells expressing EGFPWIPI2SLoop also showed conspicuous long tubules, which were more readily detectable in live cell imaging (Figure S4) than in fixed cells (Figure 4).

Role of the amphipathic α-helix of WIPI2 in β1-Integrin sorting.

A. The amphipathic α-helix. Helical wheel projections showing the CD-loop on blade 6 of the wildtype sequence, WIPI2WT, and WIPI2Sloop. Coloured arrows indicate the two pairs of amino acids that have been swapped in WIPI2Sloop. The magnitude and direction of the hydrophobic moment of the helices was predicted using the online tool Heliquest (Gautier et al., 2008). It is indicated by the vector in the centre of the wheels. Sequences of the hydrophobic loop region of WIPI2WT and WIPI2Sloop are shown, and predicted α-helices are plotted in magenta. The two pairs of hydrophobic/hydrophilic amino acids that are swapped in WIPI2Sloop are highlighted by rectangles in the sequences. B.C. β1-Integrin localisation. WIPI2KD (B) and control (C) cells were transfected with a plasmid carrying siRNA-resistant EGFPWIPI2WT or EGFPWIPI2Sloop. After 18h of viral transfection, cells were fixed, permeabilized with 0.05% saponin and stained with DAPI and antibodies to β1-Integrin. Dashed lines indicate the circumference of untransfected cells, while transfected cells are indicated by asterisks. Scale bars: 10 μm. D. Quantification of β1-Integrin immunofluorescence in cells from B and C. Regions of interest (ROIs) corresponding to cells expressing the indicated WIPI2 variants, and some regions outside the cells (background), were manually defined using Image J software. Total cell fluorescence was integrated and the background fluorescence was subtracted. The resulting total cell fluorescence was divided by the area of the cell. This value is shown in the graph. 105 cells per condition, stemming from three independent biological experiments, were quantified. Values of individual cells and means are presented by smaller and larger circles, respectively, coloured according to the independent experiment that generated them. P-values were calculated applying an unpaired Student’s t-test with unequal variances. The analysis was performed with 99% confidence: ****P < 0.00001. NS: not significant (P>0.05)

Alignments of Atg18 and WIPI1/2 orthologs have revealed a conserved FSS/TS motif in blade 2 (Courtellemont et al., 2022; Gopaldass et al., 2017; Gopaldass and Mayer, 2024; Strong et al., 2021). On the β-propeller structure of these PROPPINS, this motif is located at the membrane-distal side. It is a potential phosphorylation site and required for the interaction of Retromer with WIPI1 and Atg18 (Courtellemont et al., 2022; Feng et al., 2015). Since the overall structural arrangement of Retriever resembles that of Retromer (Bartuzi et al., 2016; Boesch et al., 2023a; Healy et al., 2023; McNally et al., 2017), we tested whether WIPI2 requires its FSSS motif to bind Retriever. To this end, we generated WIPI2S67E and WIPI2S67A to substitute the central serine in the FSSS motif and mimic effects of potential phosphorylation at this site.

The functionality of these WIPI2 variants was tested by rescuing WIPI2KD cells through expression of siRNA-resistant variants of EGFPWIPI2. Both EGFPWIPI2S67E and EGFPWIPI2S67A were expressed to similar levels as an EGFPWIPI2 wildtype allele (Figure S3). Immunofluorescence staining of non-permeabilized WIPI2KD cells showed that the expression of EGFPWIPI2WT re-established the exposure of β1-Integrin at the cell surface, while EGFPWIPI2S67A and EGFPWIPI2S67E were ineffective (Figure 5 A-C). Expression of the two variants also impaired surface exposure of β1-Integrin in control cells (Figure 5 B-D). This dominant negative effect is significant because it implies that the two variants compete with the endogenous protein. To be able to do so they should be folded, suggesting that the S67 substitutions do not grossly perturb WIPI2 structure. A conspicuous consequence of both FSSS substitutions was the accumulation of EGFPWIPI2S67A and EGFPWIPI2S67E in long tubules. These tubules were very extensive in live cell imaging and easily decayed upon fixation. (Figures. 5 and S4). Co-expression of mCherryRAB4 or mCherryRab5 revealed that these tubules often connected larger dot-like structures, which were positive for mCherryRab5 and/or mCherryRAB4 (Figure S4 A). The tubules themselves preferentially accumulated mCherryRAB4 (70% colocalization) and only little mCherryRab5 (21%) (Figure S4 B). These results are consistent with the view that WIPI2-associated tubules originate mainly from early and recycling endosomes (Figure S4) and that WIPI2 contributes fission activity to Retriever-mediated trafficking through its amphipathic helix and its FSSS motif.

Effects of the WIPI2 FSSS motif on β1-Integrin recycling.

A.B. Influence of WIPI2 variants on β1-Integrin. WIPI2-knockdown (WIPI2KD, A) and control (B) HK2 cells were transfected with a plasmid expressing siRNA-resistant wildtype or the indicated FSSS variants of EGFPWIPI2. After 18 h of viral transfection, cells were fixed (without detergent permeabilization) and stained with DAPI and antibodies to β1-Integrin (red). Dashed lines indicate the circumference of untransfected cells, while transfected cells are indicated by asterisks. Scale bars: 10 μm. C. D . Quantification of β1-Integrin immunofluorescence in cells from A and B was performed as in Figure 4 D. 105 cells per condition, stemming from three independent biological experiments, were quantified. Individual cells and means are presented by smaller and larger circles, respectively, coloured according to the independent experiment that they stem from. P-values were calculated applying an unpaired Student’s t-test with unequal variances. The analysis was performed with 99% confidence: ****P < 0.00001.

WIPI2 interacts with Retriever, Commander and SNX17

Next, we tested whether WIPI2 and WIPI1 interact with Retriever or Retromer and could become part of their respective coats. In that case, WIPI2 should interact not only with Retriever itself, but also with the sorting nexin SNX17, which binds Retriever, and perhaps the CCC subunits, which integrate Retriever into the Commander complex (Butkovič et al., 2024; Martín-González et al., 2024; Singla et al., 2024). We assayed this by co-immunoadsorption. An HA-tagged allele (WIPI2HA) was expressed in HK2 cells using lentiviruses. Extracted proteins were adsorbed to anti-HA beads and analysed by Western blotting (Figure 6). Whereas the Commander subunit CCDC93 and the Retriever subunit VPS26C co-adsorbed to the beads in extracts from cells expressing WIPI2HA (Figure 6 A, C), these signals were absent from cells expressing only non-tagged WIPI2. SNX17 also co-adsorbed with WIPI2HA. This interaction was abolished by knockdown of the Retriever subunit VPS26C (Figure 6 B), suggesting that Retriever links WIPI2 to SNX17. WIPI2 differentiated strongly between Retromer and Retriever. Although VPS26C is quite similar to the Retromer subunit VPS26, only VPS26C co-adsorbed with WIPI2HA whereas VPS26 was not detectable (Figure 6 C, D).

Integration of WIPI2 with coat subunits.

A. Interaction of WIPI2HA with the CCC complex subunit CCDC93. HK2 cells and HK2 cells stably expressing WIPI2HA were detergent solubilised, the total cell extracts were incubated with anti-HA beads and washed. Adsorbed protein was analysed by SDS-PAGE and Western blotting using the indicated antibodies. The intensity of the interacting CCDC93 was quantified with a LICOR Odyssey fluorescence imager. The background from the corresponding position in the sample from cells without HA-tag was subtracted. The resulting intensity is shown relative to the intensity of WIPI2HA signal on the beads. N = 3 biological replicates from independent experiments. Red bars show the means. Error bars represent the SEM. P values were calculated applying an unpaired Student’s t-test with unequal variances. *P < 0.01. B. Retriever-dependent interaction of WIPI2HA and SNX17. HK2 cells stably expressing WIPI2HA were treated with siRNA to VPS26C (VPS26CKD), or with non-specific siRNA (VPS26CCT), and lysed. The total cell extracts were incubated with anti-HA beads and adsorbed proteins were analysed using the indicated antibodies as in A. Values shown stem from 3 independent experiments. C. D. Selectivity for Retriever versus Retromer. The co-immunoadsorption experiments were performed as in A and analysed for co-adsorbed (C) VPS26C (Retriever) or (D) VPS26 (Retromer). N=3 independent experiments. *P < 0.01.**P < 0.001. NS=not significant.

The situation was inversed when the HA tag was attached to WIPI1. In this case, VPS26 was recovered on the beads but VPS26C was not (Figure 7A, B). SNX27, a sorting nexin that interacts with Retromer (Simonetti et al., 2022; Temkin et al., 2011), was also co-adsorbed with WIPI1HA (Figure 7C). The immunoprecipitation results were supported by parallel immunofluorescence studies. WIPI2 colocalized preferentially with Retriever and SNX17 (Figure S5) while WIPI1 colocalized preferentially with SNX27 and Retromer (Figure S6). Moreover, WIPI2 colocalized with integrin-β1 on early endosomal compartments positive also for SNX17, but not on late endosomal LAMP1 containing compartments (Figure S7). These results suggest that Retriever bridges WIPI2 and SNX17 and that Retromer bridges WIPI1 with SNX27, integrating these PROPPINs into their cognate Retriever- or Retromer- coats.

Interaction of WIPI1 with VPS26 and SNX27, but not VPS26C

A. Interaction of WIPI1HA with VPS26. HK2 cells and HK2 cells stably expressing WIPI1HA were detergent-solubilised, the total cell extracts were incubated with anti-HA beads and washed. Adsorbed protein was analysed by SDS-PAGE and Western blotting using the indicated antibodies. The intensity of the interacting VPS26 was quantified with a LICOR Odyssey fluorescence imager. The background from the corresponding position in the sample from cells without HA-tag was subtracted. The resulting intensity is shown relative to the intensity of the WIPI1HA signal on the beads. N = 3 biological replicates from independent experiments. Red bars show the means and error bars represent the SEM. P values were calculated applying an unpaired Student’s t-test with unequal variances. B. Lack of interaction of WIPI1HA with VPS26C. HK2 cells stably expressing WIPI1HA were used for co-immunoadsorption experiments as in A and decorated with the indicated antibodies. N = 3. C. Interaction of WIPI1HA with SNX27. HK2 cells stably expressing WIPI1HA were used for co-immunoadsorption experiments as in A and decorated with the indicated antibodies. N = 3. **P < 0.001. NS: not significant.

Since the FSSS motif of WIPI2 is necessary for its transport functions, we tested whether this motif affects the interactions of WIPI2 with Retriever. To this end, wildtype and the two FSSS variants of WIPI2HA were stably expressed and cell extracts were probed by the same co-adsorption approach as above (Figure 8). Whereas the Retriever subunit VPS26C was efficiently co-adsorbed with WIPI2WT-HA, WIPI2S67A-HA and WIPI2S67E-HA labilized this interaction. Cells carrying only the non-tagged, endogenous WIPI2 yielded no VPS26C signal. Thus, the FSSS motif is required for the interaction between WIPI2 and Retriever. These results were also confirmed by immunofluorescence analysis (Figure S8), where VPS26C colocalized less with WIPI2S67A and WIPI2S67E (16%) than with the wildtype form (56%).

Impact of the FSSS motif on the WIPI2-Retriever interaction.

A. HK2 cells and HK2 cells stably expressing the indicated WIPI2HA variants were detergent solubilised. Anti-HA beads were incubated with the total cell extracts, washed, and adsorbed proteins were analysed by SDS-PAGE and Western blotting using the indicated antibodies. B. Band intensities from the blots in A were quantified with a LICOR Odyssey infrared fluorescence imager and plotted as the ratio of VPS26C over WIPI2HA. N=3 independent experiments were quantified. Red bars show the means and error bars represent the SEM. P values were calculated applying an unpaired Student’s t-test with unequal variances. **P < 0.001.

The interaction of WIPI2 and Retriever is necessary for their localization on recycling endosomes

To test for the in vivo consequences of the WIPI2-Retriever interaction, we analysed the effect of the FSSS motif on the colocalization of the respective WIPI2EGFP variants with different endosomal RAB-GTPases, which were N-terminally tagged with mCherry. In line with previous studies (Polson et al., 2010; Puri et al., 2018) EGFPWIPI2WT mostly co-localized with mCherryRAB11 (58%) and mCherryRAB5 (44%) and much less with and mCherryRAB7 (6%) (Figures 9 and S9-S11). By contrast, EGFPWIPI2S67A and EGFPWIPI2S67E co-localized mainly with mCherryRAB5 (60-63%) and mCherryRAB7 (36-38%), and much less with mCherryRAB11 (12-14%). The Retriever subunit VPS35L was detected by immuno-staining. In cells expressing WIPI2WT, it mainly co-localized with mCherryRAB11 (42%) and mCherryRAB5 (43%) and less with mCherryRAB7 (23%) (Figures 9 and S9-S11). In EGFPWIPI2S67E and EGFPWIPI2S67A expressing cells, VPS35L distribution shifted, similarly as for EGFPWIPI2, indicating reduced presence on mCherryRAB11 compartments (12-13%), whereas co-localisation with mCherryRAB5 (42-44%) was maintained and that with mCherryRAB7 (26-27%) slightly increased. Collectively, these results suggest that the FSSS motif in blade 2 of WIPI2 allows this PROPPIN to associate with Retriever, and that this association is necessary for localising WIPI2 and Retriever on Rab11 containing recycling endosmes.

Role of the WIPI2 FSSS motif for recruiting WIPI2 and Retriever to Rab11 endosomes.

WIPI2-knockdown HK2 cells were transfected with plasmids expressing the indicated siRNA-resistant EGFPWIPI2 variants and mCherry-RAB5, mCherry-RAB7, or DsRed-Rab11. Cells were fixed, permeabilized, immuno-stained for VPS35L and analyzed by confocal microscopy. Examples of the quantified images are presented in Supplementary Figures S9 to S11. Colocalization with EGFPWIPI2 and VPS35L was assessed for: A. RAB11 B. RAB5 C. RAB7. Colocalization was quantified using Manders’ colocalization coefficient M2, calculated in ImageJ. M2 refers to the fraction of VPS35L colocalising with the different RAB-proteins. For the triple colocalizations WIPI2/VPS35L/RAB, M2 indicates the fraction of green pixels (WIPI2) overlapping with the pixels positive for the VPS35L/RAB colocalization. Colocalization was quantified from three independent biological experiments in a total of 120 cells. Individual cells. Individual cells and means are presented by smaller and larger circles, respectively, coloured according to the independent experiment that they stem from. P-values were calculated by an unpaired Student’s t-test with unequal variances. The analysis was performed with 99% confidence. **P < 0.001; ***P < 0.0001. NS = not significant (P>0.05).

Discussion

Our experiments reveal pathway specificity for WIPI1 and WIPI2. The two proteins support distinct cargoes and transport routes and interact with different partners.

This isoform selectivity contrasts with the roles of these proteins in their second functional domain, autophagy, which strongly depends on WIPI2 but is also significantly reduced by ablation of WIPI1 (DeLeo et al., 2021; Dooley et al., 2014; Guan et al., 2001; Obara et al., 2008; Polson et al., 2010; Proikas-Cezanne et al., 2015, 2004). Furthermore, the autophagic function of the PROPPINs also exploits other molecular features than endosomal protein exit. For autophagy, WIPI1 and its yeast orthologue Atg18 do not require the amphipathic helix in CD loop 6, which is essential for endosomal transport. It was hence proposed that WIPI1 and Atg18 perform distinct functions in autophagy and in endosomal protein exit (Courtellemont et al., 2022; DeLeo et al., 2021; Gopaldass et al., 2017).

In endosomal transport, WIPI1 and WIPI2 display selectivity in their partners (Fig. 10). WIPI1 interacts with Retromer and WIPI2 with Retriever, mirroring the impact of the two PROPPINs on the respective transport routes and cargos. WIPI1 and WIPI2 utilise the same molecular features to support these transport routes. Their FSSS motif is necessary for both isoforms to integrate with Retromer and Retriever, respectively. This, together with the similarity in the overall organisation of Retromer and Retriever, suggest that they might form complexes in a similar manner. In analogy to the WIPI1-Retromer complex, which was termed CROP (Courtellemont et al., 2022), we hence refer to the WIPI2-Retriever complex as CROP2.

WIPI2-dependent transport of β1-Integrin depends on the amphipathic character of the α-helix in CD loop 6. In Atg18 and WIPI1, this helix is necessary to convey membrane fission activity, which is used both for division of the lysosome-like vacuoles of yeast and for detachment of endosomal carriers (DeLeo et al., 2021; Gopaldass et al., 2017; Zieger and Mayer, 2012). This fission activity is potentiated by integrating Atg18 and WIPI1 with Retromer into the CROP complex (Courtellemont et al., 2022). Our observations suggest that WIPI2 contributes pathway-specific membrane fission activity by integrating with Retriever. In line with this, ablating the amphipathic character of its helix on CD loop 6 disrupts β1-Integrin recycling to the cell surface and leads to its accumulation in endosomal compartments. Disrupting the WIPI2-Retriver interaction by substitution of the FSSS motif has the same effect. It also leads to the accumulation of long tubular structures, which we presume to be exaggerated endosomal carriers that continue to grow but fail to detach. This phenotype is consistent with a block in the membrane fission step that is necessary to allow the endosomal carrier to depart.

Retriever couples to other proteins that are required for endosomal protein transport. These include the CCC complex, which associates with Retriever to yield the very large Commander complex (Boesch et al., 2023b; Healy et al., 2023; Laulumaa et al., 2024). How Commander is placed on the membrane is currently unknown, but the interaction between SNX17 and Retriever contributes to its recruitment (Butkovič et al., 2024; Healy et al., 2022; Singla et al., 2024; Steinberg et al., 2012) and presumably to the formation of a coat. Our pull-downs of WIPI2 detected not only the Retriever subunit VPS26C as an interactor, but also SNX17 and the CCC subunit CCDC93. This co-fractionation could be due to a direct interaction of WIPI2 with Retriever or reflect an interaction via another component of Commander. Our immunoprecipitation assays cannot distinguish and more detailed structural and interaction studies with pure compounds will be necessary to elucidate the nature of this interaction. Since Retriever is structurally similar to Retromer, WIPI2 is structurally similar to WIPI1 and Atg18, and a Retromer-Atg18 interaction could be reconstituted from pure proteins (Courtellemont et al., 2022) , we consider a direct WIPI2-Retriever interaction as the more likely scenario. But in both cases our observations suggest that WIPI2 integrates into the Retriever coat and might thus bring membrane fission activity to the place where it is required for carrier formation. Here, it could synergize with other factors that support membrane fission in endosomal transport, such as EHD1, which interacts with SNX17, and the actin-regulating WASH complex, which binds to Commander (Bartuzi et al., 2016; Dhawan et al., 2022, 2020; Phillips-Krawczak et al., 2015).

The requirement of two distinct WIPI proteins, using the same molecular features, in Retromer- and Retriever-dependent transport uncovers a novel conserved element between these two pathways. In combination with the similar overall organisation of Retromer and Retriever and their common interactions with EHD proteins (Arlt et al., 2015; Chi et al., 2014; Daumke et al., 2007; Deo et al., 2018; Gokool et al., 2007; Grant et al., 2001; Solinger et al., 2020) and WASH (Derivery et al., 2009; Gomez and Billadeau, 2009; Guo et al., 2024; Harbour et al., 2012; Jia et al., 2012; Phillips-Krawczak et al., 2015; Seaman et al., 2009), this supports the notion that these two membrane coats share similar mechanistic principles for their formation and/or detachment.

Materials and Methods

Antibodies and reagents

All chemical reagents were from Sigma-Aldrich unless specified otherwise. We used anti-LAMP1 (H4A3, US Biological, Life Sciences); anti- EEA1 (#2411, Cell Signaling); anti-EGFR (sc-373746, clone A-10 Santa Cruz Biotechnology; PA5-85089 ThermoFisher Scientific); anti-Integrin β1 ( ab150361, Abcam); anti-GLUT1 (ab15309, Abcam); anti-c16orf62/VPS35L (PA5-28553, ThermoFisher Scientific); anti-SNX17(HPA043867, Atlas Antibodies); anti-SNX27 (ab77799, Abcam); anti CCDC93 (20861-1AP, Proteintech); anti-α-Tubulin (T9026, Sigma); anti-Vps26 (ab181352, Abcam); anti-VPS26C/DSCR3 (ABN87, Merck Millipore), WIPI1 (W2394, Sigma); WIPI2 (HPA019852, Sigma).

As secondary antibody we used: Cy3-conjugated Affinity Pure Donkey anti-Mouse IgG H + L (Jackson Immuno Research); Cy3-conjugated AffiniPure Donkey anti-Rabbit IgG H + L (Jackson Immuno Research); Alexa Fluor®488-conjugated AffiniPure Donkey anti-Rabbit IgG H + L (Jackson Immuno Research); Alexa Fluor®488-conjugated AffiniPure Donkey anti-Mouse IgG H + L (Jackson Immuno Research); Alexa Fluor® 647-conjugated AffiniPure donkey anti-rabbit IgG H + L (Jackson Immuno Research); Alexa Fluor® 647-conjugated AffiniPure donkey anti-Mouse IgG H + L (Jackson Immuno Research). Antibodies were used at dilutions summarized in Supplementary Table1.

Other reagents: Opti-MEM (Thermo Fischer, 11058021) and Trypsin (Thermo Fischer, 27250018).

Protease inhibitor (PI) cocktail final concentrations in samples: 100 μM pefablock SC (Merck, 11429876001), 2 μM leupeptin (Merck, 11529048001), 50 μM o-phenanthroline (Merck ,131377), 0.7 μM pepstatin A (Merck, 11524488001) dissolved in methanol.

Phosphatase inhibitor cocktail final concentrations: 10 mM NaF (S-7920, Sigma); 1mM Na Orthovanadate (S6508, Sigma); 2 mM β-glycerophosphate (A2253, AppliChem) all dissolved in water.

Complementary DNA (cDNA) constructs

Vectors expressing tagged RAB proteins were purchased from Addgene: RFP-RAB4 (79,800; deposited by J.D. Johnson). DsRed-RAB11 (12679; deposited by R. Pagano); mCherry-RAB7 (61804; deposited by G. Voeltz); mCherry-RAB5 (49201; deposited by G. Voeltz). The following vectors were kindly provided by colleagues: EGFP-WIPI2 (pAR31CD vector) (Tassula Proikas-Cezanne, Tübingen, Germany). The sequences of WIPI1 and WIPI2 that have been used are given in Supplementary File 1.

Site-directed mutagenesis

All constructs were verified by DNA sequencing. EGFPWIPI2 or WIPI2HA was used as DNA template for site-directed mutagenesis (QuikChange mutagenesis system, Agilent Technologies, 200524) to generate point mutations in the FSSS motif (S67A and S67E) following the manufacturer’s protocol.

To get WIPI2S67E we used the following primers: Fw: AGATTGTTCTCCGAGAGCCTAGTGGCC and Rv GGC CAC TAG GCT CTC GGA

GAA CAA TCT; while for WIPI2S67A we used Fw: AGATTGTTCTCCGCTAGCCTAGTGGCC

Rv: GGCCACTAGGCTAGCGGAGAACAATCT (Microsynth).

WIPI2SLoop was generated by four consecutive amplifications using the following primers:

1) Fw: TGGACC GGGTACAAAGGGAAAGTG CTC

Rv: GAGCACTTTCCCTTTGTACCCGGTCCA;

2) Fw: GGGTACAAAGGGTTCGTGCTCATGGCC

Rv: GGCCATGAGCACGAACCCTTTGTACCC;

3) Fw: CTC ATG GCC TCC TAC AGC TAC CTG CCT

Rv: AGGCAGGTAGCTGTAGGAGGCCATGAG;

4) Fw: GCC TCC TAC AGC ACC CTG CCT TCC CAA

Rv: TTGGGAAGGCAGGGTGCTGTAGGAGGC

Non-mutated template vector was removed from the PCR mixture through digestion by the enzyme Dpn1 for 1 h at 37°C. The product was purified using NucleoSpin PCR Clean-up (Macherey-Nagel, 740609.50S) and transformed into Escherichia coli. Plasmid DNA was purified and sequenced. The site-directed mutagenesis system was also used to generate the siRNA-resistant cDNA for WIPI2 with primers Fw: CGATAGTCCTTTAGCCGCA and Rv: TGCGGCTAAAGGACTATCG. All oligonucleotides were synthesized by Microsynth.

Cell culture, transfection and treatments

HK2 cells were grown in DMEM-HAM’s F12 (Thermo Fischer, 11765054) supplemented with 5% fetal calf serum (Gibco, 10270106), 50 U/mL penicillin/50 mg/mL streptomycin (ThermoFischer, 15140148), ITS (5 μg/mL insulin, 5 μg/mL TF, 5 ng/mL selenium; LuBio Science, 00-101-100ML). Cells were grown at 37°C in 5% CO2 and at 98% humidity.

HK2 cells were transfected with plasmids using the X-tremeGENETM HP DNA transfection reagent (Sigma-Aldrich, 6366546001), unless otherwise specified, according to the manufacturer’s instructions, and incubated for 18-24 h before fixation or live cell imaging. The HK2 cell line was checked for mycoplasma contamination by a PCR-based method. All cell-based experiments were independently repeated at least three times.

Generation of HK2 cells expressing WIPI2HA

HK2 cell lines stably expressing the three forms (WT, S67A, S67E) of C-terminal HA-tagged WIPI2 were generated by lentiviral transduction. WIPI2 was cloned into modified pLKO.1 lentiviral vector (available from Sigma) between Xba1 and EcoR1 restriction sites. To obtain WIPI2HA a sequence of 3Gly - 1Ser - 3Gly - 1Ser - HA - HA (designed and purchased from GeneScript) was added using Gibson assembly according to the manufacturer’s protocol. Finally, the lentiviral plasmid DNA was purified and sequenced. Lentiviral transfer vector together with third generation envelope and packaging plasmids were transfected into 293T cells to generate lentivirus. Packaged lentivirus containing supernatant was added to recipient HK2 cells in the presence of 10 μg/ml Polybrene (AL-118, Sigma-Aldrich). HK2 cells were stably infected with a lentiviral vector for constitutive WIPI2HA expression for 6 h. 2 days after infection, cells were selected with 1 μg/ml of puromycin for 3 days. Protein samples were collected and used for IP experiments 7 days after infection.

RNA interference

HK2 cells were transfected with siRNA for 72 h using Lipofectamine® RNAiMax (Thermo Fisher Scientific, 13778150) according to the manufacturer’s instructions. Control cells were treated with identical concentrations of siGENOME Control Pool Non-Targeting from Dharmacon (D-001206-13-05). siRNAs targeting WIPI2 were from Dharmacon (ON-TARGETplus Human WIPI2 J-020521). siRNAs were used at a final concentration of 20 nM.

Endocytosis assays

EGFR degradation

HK2 cells were seeded in DMEM-HAM’s F12 supplemented with 5% foetal calf serum, 50 U/mL penicillin/50 mg/mL streptomycin and ITS (5 μg/mL insulin, 5 μg/mL TF, 5 ng/mL selenium. The day after the growth medium was replaced with the same but without serum for 24 h and HK2 cells were stimulated with 100 ng/ml EGF for the indicated periods of time. The cells were fixed for 10 min in 4% paraformaldehyde in PBS at different time points after stimulation for IF studies (Figure 1 A-B) or washed and lysed for immunoblotting analysis (Figure D)

Immunofluorescence

Cells were fixed for 10 min in 4% paraformaldehyde in PBS. After fixation, cells were permeabilized in 0.05% (w:v) saponin (Sigma-Aldrich, 558255), 0.5% (w:v) BSA and 50 mM NH4Cl in PBS (blocking buffer) for 30 min at room temperature. The cells were incubated for 1 h with primary antibodies in blocking buffer, washed three times in PBS, incubated for 1 h with the secondary antibodies (Alexa Fluor-conjugated), washed three times in PBS, mounted with fluorescence mounting medium (Dako, S3023) on slides and analysed by confocal microscopy.

Confocal fluorescence microscopy, image processing, and colocalization analysis

HK2 cells were grown to 70% confluence on glass coverslips and immunofluorescence microscopy was performed as described above. The experiments were repeated at least three times and representative images are shown. Colocalization was analysed by acquiring serial sections from about 100 cells per sample. Images were exported in TIFF format and processed as previously described (Vicinanza et al., 2011). Images of samples to be compared were acquired using the same settings (i.e. laser power, photomultiplier gain and pinhole size), avoiding pixel saturation. The images were processed in the same way using ImageJ software. Channels from each image were converted into 8-bit format and the ‘Auto Local Threshold’ Plug-in with the ‘Default’ method was used to segment grayscale images, identify the structures of interest and subtract background.

GLUT1, β1-Integrin, EEA1, and LAMP1 fluorescence levels were quantified using ImageJ. z-stack images were acquired and compressed into a single plane using the ‘maximum intensity Z-projection’ function in Image J. Individual cells were selected using the freeform drawing tool to create a ROI (ROI). The ‘Measure’ function provided the area, the mean grey value and integrated intensity of the ROI. The mean background level was obtained by measuring the intensity in three different regions outside the cells, dividing them by the area of the regions measured, and averaging the values obtained. The background was subtracted from each cell to obtain the CTCF (corrected total cell fluorescence). It was calculated using the formula: CTCF=integrated intensity of cell ROI − (area of ROI × mean fluorescence of background).

To quantify the degree of colocalization, confocal z-stacks were acquired, but a single panel was used for the measurement. Single channels from each image in 8-bit format were thresholded to subtract background and then the "Just Another Colocalisation Plug-in" (JACoP) ImageJ software was used to measure the overlap coefficient according to Manders. Manders’ coefficient indicates the overlap of the signals and represents the degree of colocalization: R=∑iS1i⋅S2i∑i(S1i)2⋅∑i(S21)2 where S1 represents signal intensity of pixels in the channel 1 and S2 represents signal intensity of pixels in the channel 2. The colocalization coefficients M1 and M2 indicate the contribution of each channel to the pixels. They are independent of the intensity of the overlapping pixels, but they are sensitive to background, so a threshold must be set (Manders et al., 2011).

For the triple colocalization analysis (Figures S9-S11) we first performed colocalization between two single channels (i.e., red and blue) and then took that subset of overlapping spots (either a binary image or a list of spots) to measure how many colocalized with green signals.

To measure colocalization of RAB-proteins on tubules in cells overexpressing EGFPWIPI2 variants (Figure S4) we used an approach based on the detection of signal edges through the “Canny or Sobel filters” Plug-in for ImageJ (Canny, 1986) . The surrounded fields were filled to generate binary mask images. In this way, we were able to select tubules in order to separate signal from background, but also to determine a common region for analyzing both channels (i.e., red for Rabs and green for WIPI2). These mask images were then subjected to thresholding as described above and colocalization was evaluated with the ImageJ JACOP Plug-in by calculating Manders’ coefficient. Confocal microscopy was performed on an inverted confocal laser microscope (Zeiss LSM 880 with airyscan) with a 63x 1.4 NA oil immersion lens unless stated otherwise.

Gel electrophoresis and Western blot

Control cells were plated into 12-well tissue culture test plates (TPP) until they reached around 80-90% confluency, except for knockdown-cells, which were cultured until 72h after transfection with the siRNAs by reaching a confluence of 80%. Cells were then washed three times with ice-cold PBS (phosphate-buffered saline), scraped, and proteins were extracted in ice-cold lysis buffer (150 mM NaCl, 2 mM EDTA, 40 mM HEPES-NaOH pH 7.4, and 1% Triton X-100 supplemented with phosphatases and protease inhibitor cocktail (see details in Reagent section). Protein extracts were supplemented with 1/4 volume of 5x reducing sample buffer (250 mM Tris-Cl, pH 6.8, 5% β-mercaptoethanol, 10% SDS, 30% glycerol, 0.02% bromophenol blue) and heated to 95 °C for 5 min. The samples were run on either 8%, 10% or 12.5% SDS-polyacrylamide gels (W x L x H: 8.6 x 6.8 x 0.15 cm). The stacking gels were prepared as follows: 6% acrylamide, 0.16% bis-acrylamide, 0.1 M Tris, pH 6.8, 0.1% SDS, 0.1% TEMED, 0.05% ammonium persulfate. Running gels were: 10% or 12.5% acrylamide, 0.27% or 0.34% bis-acrylamide, 0.38 M Tris, pH 8.8, 0.1% SDS (Applichem, 475904-M), 0.06% TEMED (Applichem, A1148), 0.06% APS (Applichem, A2941). The gels were run at constant current (20–30 mA). Proteins were blotted onto nitrocellulose membrane by the semi-dry method for 80 min at 400 mA (Trans-Blot® SD Semi-Dry Electrophoretic Transfer Cell, Bio-Rad). After incubation with the primary antibody, signals were detected by secondary antibodies coupled to infrared dyes (LI-COR) and detected on a LI-COR Odyssey infrared fluorescence imager. Images were exported as TIFF files and processed in Adobe Photoshop. Band intensity was quantified using ImageJ band analysis.

Immunoprecipitation of Retriever

HK2 cells expressing the three variants (WT, S67E, S67A) of WIPI2 tagged with HA were grown to 80% confluence in 15 cm dishes. After rinsing the cells three times with phosphate buffered saline (PBS 1X), they were detached and lysed by incubating 15 min at 4°C with lysis buffer: 150 mM NaCl, 2 mM EDTA, 40 mM HEPES-NaOH pH 7.4, 1% Triton X-100, supplemented with phosphatases and proteases inhibitor cocktails (details in Reagent section) and 1 mM PMSF out of a 200 mM solution in ethanol, which was freshly prepared immediately before use. The lysate was spun for 10 min at 10,000 x g on a cold bench-top centrifuge). 2.5% of the clarified extracts was reserved as “input”. Each clarified extract was incubated for 2 h at 4°C with 20 μl of Anti-HA High Affinity antibody (clone 3F10) (Roche #118150116001). Then, samples were washed three times with lysis buffer and, after discarding the last wash, the beads were resuspended in 50 μl of pre-warmed elution buffer (2% SDS, 1mM EDTA, 10mM DTT, 20mM Tris HCl pH 7.4, 100 µM PMSF) and shaken at 55°C for 10 min. Samples were centrifuged, the supernatants were transferred to new Eppendorf tubes, and 5x sample buffer containing 100 mM of fresh DTT was added. Samples were kept at 95°C for 5 min and centrifuged in a table to centrifuge. All supernatants were transferred to a new Eppendorf tube yielding the “IP” samples. Inputs and IP were loaded on 8% or 10% SDS–polyacrylamide gels.

Statistical analysis

Differences between the means have been evaluated by an unpaired Student’s t-test assuming unequal variances unless otherwise specified. They are considered as significant differences when p < 0.01. Experiments were independently repeated at least three times. For fluorescence microscopy quantifications at least 30 cells were analysed from each experiment. Representative images are shown. Western blotting experiments were independently repeated at least three times and representative blots are shown. Means ± standard deviation (s.d.) are shown unless specified otherwise.